% SOURCE: https://en.wikisource.org/wiki/Shakespeare_-_First_Folio_facsimile_(1910)/The_Tragedy_of_Hamlet (2022-07-27)

\documentclass[a5paper,DIV=calc,11pt]{scrbook}
\usepackage[british]{babel}
\usepackage{verse}
\usepackage{dramatist}
\usepackage{microtype}
\usepackage{lettrine}
\usepackage[defaultsans]{cantarell}

\newcommand{\updatemarks}{\markboth{Act \Roman{act}}{Scene \Roman{act}--\Roman{scene}}}


\begin{document}
\title{The Tragedie of Hamlet}
\author{William Shakespeare}
\date{1601}
\maketitle

\tableofcontents

\Character[Hamlet – son of the late king and nephew of the present king, Claudius]{Hamlet}{ham}
\Character[Claudius – king of Denmark, Hamlet's uncle and brother to the former king]{Claudius}{king}
\Character[Gertrude – queen of Denmark and Hamlet's mother]{Gertrude}{queen}
\Character[Polonius – chief counsellor to the king]{Polonius}{pol}
\Character[Ophelia – Polonius's daughter]{Ophelia}{ophe}
\Character[Horatio – friend of Hamlet]{Horatio}{hor}
\Character[Laertes – Polonius's son]{Laertes}{laer}
\begin{CharacterGroup}{courtiers}
    \GCharacter{Voltimand}{Voltimand}{volt}
    \GCharacter{Cornelius}{Cornelius}{corn}
\end{CharacterGroup}
\begin{CharacterGroup}{courtiers, friends of Hamlet}
    \GCharacter{Rosencrantz}{Rosencrantz}{rosin}
    \GCharacter{Guildenstern}{Guildenstern}{guil}
\end{CharacterGroup}
\Character[Osricke – a courtier]{Osricke}{osr}
\begin{CharacterGroup}{officers}
    \GCharacter{Marcellus}{Marcellus}{mar}
    \GCharacter{Barnardo}{Barnardo}{barn}
\end{CharacterGroup}
\Character[Francisco – a soldier]{Francisco}{fran}
\Character[Reynaldo – Polonius's servant]{Raynaldo}{reyn}
\Character[Ghost – the ghost of Hamlet's father]{Ghost}{gho}
\Character[Fortinbras – prince of Norway]{Fortinbras}{for}
%\Character[Gravediggers – a pair of sextons]{Gravediggers}{dig}
\begin{CharacterGroup}{theatre group}
    \GCharacter{Player King}{P. King}{playking}
    \GCharacter{Player Queen}{P. Queen}{playqueen}
    \GCharacter{Lucianus}{Lucianus}{playlucianus}
\end{CharacterGroup}

\DramPer

\chaptermark{Dramatis Personæ}

% ---

\setlength\vindent{1em}

\act

\scene

\updatemarks

\markboth{Act \Roman{act}}{Scene \Roman{act}--\Roman{scene}}

\StageDir{Enter \barn and \fran two Centinels.}

\begin{drama*}
    \barnspeaks Ho's there?

    \franspeaks Nay answer me: Stand \& vnfold your selfe.

    \barnspeaks Long liue the King.
    
    \franspeaks \barn?

    \barnspeaks He.

    \franspeaks You come most carefully vpon your houre.

    \barnspeaks 'Tis now strook twelue, get thee to bed Francisco.

    \franspeaks For this releefe much thankes: 'Tis bitter cold,\\
    And I am sicke at heart.

    \barnspeaks Haue you had quiet Guard?

    \franspeaks Not a Mouse stirring.

    \barnspeaks Well, goodnight. If you do meet \hor and \mar Marcellus, the Riuals of my Watch, bid them make hast.

    \direct{Enter \hor and \mar.}

    \franspeaks I thinke I heare them. Stand: who's there?

    \horspeaks Friends to this ground.

    \marspeaks And Leige-men to the Dane.

    \franspeaks Giue you good night.

    \marspeaks O farwel honest Soldier, who hath relieu'd you?

    \franspeaks \barn ha's my place: giue you goodnight.\\(\textit{Exit \fran.})

    \marspeaks Holla \barn.

    \barnspeaks Say, what is \hor there?

    \horspeaks A peece of him.
    
    \barnspeaks Welcome \hor, welcome good \mar.

    \marspeaks What, ha's this thing appear'd againe to night.

    \barnspeaks I haue seene nothing.

    \marspeaks \hor saies, 'tis but our Fantasie,\\
    And will not let beleefe take hold of him\\
    Touching this dreaded sight, twice seene of vs,\\
    Therefore I haue intreated him along\\
    With vs, to watch the minutes of this Night,\\
    That if againe this Apparition come,\\
    He may approue our eyes, and speake to it.

    \horspeaks Tush, tush, 'twill not appeare.

    \barnspeaks Sit downe a-while,\\
    And let vs once againe assaile your eares,\\
    That are so fortified against our Story,\\
    What we two Nights haue seene.

    \horspeaks Well, sit we downe,\\
    And let vs heare \barn speake of this.

    \barnspeaks Last night of all,\\
    When yond same Starre that's Westward from the Pole\\
    Had made his course t'illume that part of Heauen\\
    Where now it burnes, Marcellus and my selfe,\\
    The Bell then beating one.

    \marspeaks Peace, breake thee of: \hfill(\textit{Enter the \gho}.)\\
    Looke where it comes againe.

    \barnspeaks In the same figure, like the King that's dead.

    \marspeaks Thou art a Scholler; speake to it \hor.

    \barnspeaks Lookes it not like the King? Marke it \hor.

    \horspeaks Most like: It harrowes me with fear \& wonder

    \barnspeaks It would be spoke too.

    \marspeaks Question it Horatio.

    \horspeaks What art thou that vsurp'st this time of night,\\
    Together with that Faire and Warlike forme\\
    In which the Maiesty of buried Denmarke\\
    Did sometimes march: By Heauen I charge thee speake.

    \marspeaks It is offended.

    \barnspeaks See, it stalkes away.

    \horspeaks Stay: speake; speake: I Charge thee, speake. \hfill(\textit{Exit the \gho.})

    \marspeaks 'Tis gone, and will not answer.

    \marspeaks How now \hor? You tremble \& look pale:\\
    Is not this something more then Fantasie?\\
    What thinke you on't?

    \horspeaks Before my God, I might not this beleeue\\
    Without the sensible and true auouch\\
    Of mine owne eyes.

    \marspeaks Is it not like the King?

    \horspeaks As thou art to thy selfe,\\
    Such was the very Armour he had on,\\
    When th'Ambitious Norwey combatted:\\
    So frown'd he once, when in an angry parle\\
    He smot the sledded Pollax on the Ice.\\
    'Tis strange.

    \marspeaks Thus twice before, and iust at this dead houre,\\
    With Martiall stalke, hath he gone by our Watch.

    \horspeaks In what particular thought to work, I know not:\\
    But in the grosse and scope of my Opinion,\\
    This boades some strange erruption to our State.

    \marspeaks Good now sit downe, \& tell me he that knowes\\
    Why this same strict and most obseruant Watch,\\
    So nightly toyles the subiect of the Land,\\
    And why such dayly Cast of Brazon Cannon\\
    And Forraigne Mart for Implements of warre:\\
    Why such impresse of Ship-wrights, whose sore Taske\\
    Do's not diuide the Sunday from the weeke,\\
    What might be toward, that this sweaty hast\\
    Doth make the Night ioynt-Labourer with the day:\\
    Who is't that can informe me?
\end{drama*}

\scene

\updatemarks

\StageDir{Enter \king King of Denmarke, \queen the Queene, \ham, \pol, \laer, and his Sister \ophe, Lords Attendant.}

\begin{drama*}
    \kingspeaks Though yet of \ham our deere Brothers death\\
    The memory be greene: and that it vs befitted\\
    To beare our hearts in greefe, and our whole Kingdome\\
    To be contracted in one brow of woe:\\
    Yet so farre hath Discretion fought with Nature,\\
    That we with wisest sorrow thinke on him,\\
    Together with remembrance of our selues.\\
    Therefore our sometimes Sister, now our Queen,\\
    Th'Imperiall Ioyntresse of this warlike State,\\
    Haue we, as 'twere, with a defeated ioy,\\
    With one Auspicious, and one Dropping eye,\\
    With mirth in Funerall, and with Dirge in Marriage,\\
    In equall Scale weighing Delight and Dole\\
    Taken to Wife; nor haue we heerein barr'd\\
    Your better Wisedomes, which haue freely gone\\
    With this affaire along, for all our Thankes.\\
    Now followes, that you know young Fortinbras,\\
    Holding a weake supposall of our worth;\\
    Or thinking by our late deere Brothers death,\\
    Our State to be disioynt, and out of Frame,\\
    Colleagued with the dreame of his Aduantage;\\
    He hath not fayl'd to pester vs with Message,\\
    Importing the surrender of those Lands\\
    Lost by his Father: with all Bonds of Law\\
    To our most valiant Brother. So much for him.\\
    \textit{Enter \volt and \corn.}\\
    Now for our selfe, and for this time of meeting\\
    Thus much the businesse is. We haue heere writ\\
    To Norway, Vncle of young Fortinbras,\\
    Who Impotent and Bedrid, scarsely heares\\
    Of this his Nephewes purpose, to suppresse\\
    His further gate heerein. In that the Leuies,\\
    The Lists, and full proportions are all made\\
    Out of his subiect: and we heere dispatch\\
    You good Cornelius, and you Voltemand,\\
    For bearing of this greeting to old Norway,\\
    Giuing to you no further personall power\\
    To businesse with the King, more then the scope\\
    Of these dilated Articles allow:\\
    Farewell, and let your hast commend your duty.
    
    \voltspeaks In that, and all things, will we shew our duty.
    
    \kingspeaks We doubt it nothing, heartily farewell.\\
    \textit{Exit Voltemand and Cornelius.}\\
    And now \laer, what's the newes with you?\\
    You told vs of some suite. What is't \laer?\\
    You cannot speake of Reason to the Dane,\\
    And loose your voyce. What would'st thou beg \laer,\\
    That shall not be my Offer, not thy Asking?\\
    The Head is not more Natiue to the Heart,\\
    The Hand more instrumentall to the Mouth,\\
    Then is the Throne of Denmarke to thy Father.\\
    What would'st thou haue \laer?

    \laerspeaks Dread my Lord,\\
    Your leaue and fauour to returne to France,\\
    From whence, though willingly I came to Denmarke\\
    To shew my duty in your Coronation,\\
    Yet now I must confesse, that duty done,\\
    My thoughts and wishes bend againe towards France,\\
    And bow them to your gracious leaue and pardon.
    
    \kingspeaks Haue you your Fathers leaue?\\
    What sayes \pol?
    
    \polspeaks He hath my Lord:\\
    I do beseech you giue him leaue to go.
    
    \kingspeaks Take thy faire houre \laer, time be thine,\\
    And thy best graces spend it at thy will:\\
    But now my Cosin \ham, and my Sonne?
    
    \hamspeaks A little more then kin, and lesse then kinde.
    
    \kingspeaks How is it that the Clouds still hang on you?
    
    \hamspeaks Not so my Lord, I am too much i'th' Sun.
    
    \queenspeaks Good \ham cast thy nightly colour off,\\
    And let thine eye looke like a Friend on Denmarke.\\
    Do not for euer with thy veyled lids\\
    Seeke for thy Noble Father in the dust;\\
    Thou know'st 'tis common, all that liues must dye,\\
    Passing through Nature, to Eternity.
    
    \hamspeaks I Madam, it is common.
    
    \queenspeaks If it be;\\
    Why seemes it so particular with thee.
    
    \hamspeaks Seemes Madam? Nay, it is: I know not Seemes:\\
    'Tis not alone my Inky Cloake (good Mother)\\
    Nor Customary suites of solemne Blacke,\\
    Nor windy suspiration of forc'd breath,\\
    No, nor the fruitfull Riuer in the Eye,\\
    Nor the deiected hauiour of the Visage,\\
    Together with all Formes, Moods, shewes of Griefe,\\
    That can denote me truly. These indeed Seeme,\\
    For they are actions that a man might play:\\
    But I haue that Within, which passeth show;\\
    These, but the Trappings, and the Suites of woe.
    
    \kingspeaks 'Tis sweet and commendable\\
    In your Nature \ham,\\
    To giue these mourning duties to your Father:\\
    But you must know, your Father lost a Father,\\
    That Father lost, lost his, and the Suruiuer bound\\
    In filiall Obligation, for some terme\\
    To do obsequious Sorrow. But to perseuer\\
    In obstinate Condolement, is a course\\
    Of impious stubbornnesse. 'Tis vnmanly greefe,\\
    It shewes a will most incorrect to Heauen,\\
    A Heart vnfortified, a Minde impatient,\\
    An Vnderstanding simple, and vnschool'd:\\
    For, what we know must be, and is as common\\
    As any the most vulgar thing to sence,\\
    Why should we in our peeuish Opposition\\
    Take it to heart? Fye, 'tis a fault to Heauen,\\
    A fault against the Dead, a fault to Nature,\\
    To Reason most absurd, whose common Theame\\
    Is death of Fathers, and who still hath cried,\\
    From the first Coarse, till he that dyed to day,\\
    This must be so. We pray you throw to earth\\
    This vnpreuayling woe, and thinke of vs\\
    As of a Father; For let the world take note,\\
    You are the most immediate to our Throne,\\
    And with no lesse Nobility of Loue,\\
    Then that which deerest Father beares his Sonne,\\
    Do I impart towards you. For your intent\\
    In going backe to Schoole in Wittenberg,\\
    It is most retrograde to our desire:\\
    And we beseech you, bend you to remaine\\
    Heere in the cheere and comfort of our eye,\\
    Our cheefest Courtier Cosin, and our Sonne.

    \hamspeaks A little more then kin, and lesse then kinde.
    
    \kingspeaks How is it that the Clouds still hang on you?
    
    \hamspeaks Not so my Lord, I am too much i'th' Sun.
    
    \queenspeaks Good \ham cast thy nightly colour off,\\
    And let thine eye looke like a Friend on Denmarke.\\
    Do not for euer with thy veyled lids\\
    Seeke for thy Noble Father in the dust;\\
    Thou know'st 'tis common, all that liues must dye,\\
    Passing through Nature, to Eternity.
    
    \hamspeaks I Madam, it is common.
    
    \queenspeaks If it be;\\
    Why seemes it so particular with thee.
    
    \hamspeaks Seemes Madam? Nay, it is: I know not Seemes:\\
    'Tis not alone my Inky Cloake (good Mother)\\
    Nor Customary suites of solemne Blacke,\\
    Nor windy suspiration of forc'd breath,\\
    No, nor the fruitfull Riuer in the Eye,\\
    Nor the deiected hauiour of the Visage,\\
    Together with all Formes, Moods, shewes of Griefe,\\
    That can denote me truly. These indeed Seeme,\\
    For they are actions that a man might play:\\
    But I haue that Within, which passeth show;\\
    These, but the Trappings, and the Suites of woe.
    
    \kingspeaks 'Tis sweet and commendable\\
    In your Nature \ham,\\
    To giue these mourning duties to your Father:\\
    But you must know, your Father lost a Father,\\
    That Father lost, lost his, and the Suruiuer bound\\
    In filiall Obligation, for some terme\\
    To do obsequious Sorrow. But to perseuer\\
    In obstinate Condolement, is a course\\
    Of impious stubbornnesse. 'Tis vnmanly greefe,\\
    It shewes a will most incorrect to Heauen,\\
    A Heart vnfortified, a Minde impatient,\\
    An Vnderstanding simple, and vnschool'd:\\
    For, what we know must be, and is as common\\
    As any the most vulgar thing to sence,\\
    Why should we in our peeuish Opposition\\
    Take it to heart? Fye, 'tis a fault to Heauen,\\
    A fault against the Dead, a fault to Nature,\\
    To Reason most absurd, whose common Theame\\
    Is death of Fathers, and who still hath cried,\\
    From the first Coarse, till he that dyed to day,\\
    This must be so. We pray you throw to earth\\
    This vnpreuayling woe, and thinke of vs\\
    As of a Father; For let the world take note,\\
    You are the most immediate to our Throne,\\
    And with no lesse Nobility of Loue,\\
    Then that which deerest Father beares his Sonne,\\
    Do I impart towards you. For your intent\\
    In going backe to Schoole in Wittenberg,\\
    It is most retrograde to our desire:\\
    And we beseech you, bend you to remaine\\
    Heere in the cheere and comfort of our eye,\\
    Our cheefest Courtier Cosin, and our Sonne.
    
    \queenspeaks Let not thy Mother lose her Prayers \ham:\\
    I prythee stay with vs, go not to Wittenberg.
    
    \hamspeaks I shall in all my best\\
    Obey you Madam.
    
    \kingspeaks Why 'tis a louing, and a faire Reply,\\
    Be as our selfe in Denmarke. Madam come,\\
    This gentle and vnforc'd accord of \ham\\
    Sits smiling to my heart; in grace whereof,\\
    No iocond health that Denmarke drinkes to day,\\
    But the great Cannon to the Clowds shall tell,\\
    And the Kings Rouce, the Heauens shall bruite againe,\\
    Respeaking earthly Thunder. Come away. \hfill(\textit{Exeunt})
    
    \direct{Manet \ham.}
    
    \hamspeaks Oh that this too too solid Flesh, would melt,\\
    Thaw, and resolue it selfe into a Dew:\\
    Or that the Euerlasting had not fixt\\
    His Cannon 'gainst Selfe-slaughter. O God, O God!\\
    How weary, stale, flat, and vnprofitable\\
    Seemes to me all the vses of this world?\\
    Fie on't? Oh fie, fie, 'tis an vnweeded Garden\\
    That growes to Seed: Things rank, and grosse in Nature\\
    Possesse it meerely. That it should come to this:\\
    But two months dead: Nay, not so much; not two,\\
    So excellent a King, that was to this\\
    Hiperion to a Satyre: so louing to my Mother,\\
    That he might not beteene the windes of heauen\\
    Visit her face too roughly. Heauen and Earth\\
    Must I remember: why she would hang on him,\\
    As if encrease of Appetite had growne\\
    By what it fed on; and yet within a month?\\
    Let me not thinke on't: Frailty, thy name is woman.\\
    A little Month, or ere those shooes were old,\\
    With which she followed my poore Fathers body\\
    Like Niobe, all teares. Why she, euen she.\\
    (O Heauen! A beast that wants discourse of Reason\\
    Would haue mourn'd longer) married with mine Vnkle,\\
    My Fathers Brother: but no more like my Father,\\
    Then I to Hercules. Within a Moneth?\\
    Ere yet the salt of most vnrighteous Teares\\
    Had left the flushing of her gauled eyes,\\
    She married. O most wicked speed, to post\\
    With such dexterity to Incestuous sheets:\\
    It is not, nor it cannot come to good.\\
    But breake my heart, for I must hold my tongue.
    
    \direct{Enter \hor, \barn, and \mar.}
    
    \horspeaks Haile to your Lordship.
    
    \hamspeaks I am glad to see you well:\\
    \hor, or I do forget my selfe.
    
    \horspeaks The same my Lord,\\
    And your poore Seruant euer.
    
    \hamspeaks Sir my good friend,\\
    Ile change that name with you:\\
    And what make you from Wittenberg \hor?
    
    \direct{\mar.}
    
    \marspeaks My good Lord.
    
    \hamspeaks I am very glad to see you: good euen Sir.\\
    But what in faith make you from Wittemberge?
    
    \horspeaks A truant disposition, good my Lord.
    
    \hamspeaks I would not haue your Enemy say so;\\
    Nor shall you doe mine eare that violence,\\
    To make it truster of your owne report\\
    Against your selfe. I know you are no Truant:\\
    But what is your affaire in Elsenour?\\
    Wee'l teach you to drinke deepe, ere you depart.
    
    \horspeaks My Lord, I came to see your Fathers Funerall.
    
    \hamspeaks I pray thee doe not mock me (fellow Student)\\
    I thinke it was to see my Mothers Wedding.
    
    \horspeaks Indeed my Lord, it followed hard vpon.
    
    \hamspeaks Thrift thrift \hor: the Funerall Bakt-meats\\
    Did coldly furnish forth the Marriage Tables;\\
    Would I had met my dearest foe in heauen,\\
    Ere I had euer seene that day \hor.\\
    My father, me thinkes I see my father.
    
    \horspeaks Oh where my Lord?
    
    \hamspeaks In my minds eye (\hor)\
    
    \horspeaks I saw him once; he was a goodly King.
    
    \hamspeaks He was a man, take him for all in all:\\
    I shall not look vpon his like againe.
    
    \horspeaks My Lord, I thinke I saw him yesternight.
    
    \hamspeaks Saw? Who?
    
    \horspeaks My Lord, the King your Father.
    
    \hamspeaks The King my Father?
    
    \horspeaks Season your admiration for a while\\
    With an attent eare; till I may deliuer\\
    Vpon the witnesse of these Gentlemen,\\
    This maruell to you.
    
    \hamspeaks For Heauens loue let me heare.
    
    \horspeaks Two nights together, had these Gentlemen\\
    (\mar and \barn) on their Watch\\
    In the dead wast and middle of the night\\
    Beene thus encountred. A figure like your Father,\\
    Arm'd at all points exactly, Cap a Pe,\\
    Appeares before them, and with sollemne march\\
    Goes slow and stately: By them thrice he walkt,\\
    By their opprest and feare-surprized eyes,\\
    Within his Truncheons length; whilst they bestil'd\\
    Almost to Ielly with the Act of feare,\\
    Stand dumbe and speake not to him. This to me\\
    In dreadfull secrecie impart they did,\\
    And I with them the third Night kept the Watch,\\
    Whereas they had deliuer'd both in time,\\
    Forme of the thing; each word made true and good,\\
    The Apparition comes. I knew your Father:\\
    These hands are not more like.
    
    \hamspeaks But where was this?
    
    \marspeaks My Lord, vpon the platforme where we watcht.
    
    \hamspeaks Did you not speake to it?
    
    \horspeaks My Lord, I did;\\
    But answere made it none: yet once me thought\\
    It lifted vp it head, and did addresse\\
    It selfe to motion, like as it would speake:\\
    But euen then, the Morning Cocke crew lowd;\\
    And at the sound it shrunke in hast away,\\
    And vanisht from our sight.
    
    \hamspeaks Tis very strange.
    
    \horspeaks As I doe liue my honourd Lord 'tis true;\\
    And we did thinke it writ downe in our duty\\
    To let you know of it.
    
    \hamspeaks Indeed, indeed Sirs; but this troubles me.\\
    Hold you the watch to Night?
    
    \speaker{\mar and \barn} We doe my Lord.
    
    \hamspeaks Arm'd, say you?
    
    \speaker{\mar and \barn} Arm'd, my Lord.
    
    \hamspeaks From top to toe?
    
    \speaker{\mar and \barn} My Lord, from head to foote.
    
    \hamspeaks Then saw you not his face?
    
    \horspeaks O yes, my Lord, he wore his Beauer vp.
    
    \hamspeaks What, lookt he frowningly?
    
    \horspeaks A countenance more in sorrow then in anger.
    
    \hamspeaks Pale, or red?
    
    \horspeaks Nay very pale.
    
    \hamspeaks And fixt his eyes vpon you?
    
    \horspeaks Most constantly.
    
    \hamspeaks I would I had beene there.
    
    \horspeaks It would haue much amaz'd you.
    
    \hamspeaks Very like, very like: staid it long?
    
    \horspeaks While one with moderate hast might tell a hundred.
    
    \speaker{All} Longer, longer.
    
    \horspeaks Not when I saw't.
    
    \hamspeaks His Beard was grisly? no.
    
    \horspeaks It was, as I haue seene it in his life,\\
    A Sable Siluer'd.
    
    \hamspeaks Ile watch to Night; perchance 'twill wake againe.
    
    \horspeaks I warrant you it will.
    
    \hamspeaks If it assume my noble Fathers person,\\
    Ile speake to it, though Hell it selfe should gape\\
    And bid me hold my peace. I pray you all,\\
    If you haue hitherto conceald this sight;\\
    Let it bee treble in your silence still:\\
    And whatsoeuer els shall hap to night,\\
    Giue it an vnderstanding but no tongue;\\
    I will requite your loues; so, fare ye well:\\
    Vpon the Platforme twixt eleuen and twelue,\\
    Ile visit you.
    
    \speaker{All} Our duty to your Honour. \hfill(\textit{Exeunt.})
    
    \hamspeaks Your loue, as mine to you: farewell.\\
    My Fathers Spirit in Armes? All is not well:\\
    I doubt some foule play: would the Night were come;\\
    Till then sit still my soule; foule deeds will rise,\\
    Though all the earth orewhelm them to mens eies. \hfill(\textit{Exit})
\end{drama*}

\scene

\updatemarks

\StageDir{Enter \laer and \ophe.}

\begin{drama*}
    \laerspeaks My necessaries are imbark't; Farewell:\\
    And Sister, as the Winds giue Benefit,\\
    And Conuoy is assistant; doe not sleepe,\\
    But let me heare from you.
    
    \ophespeaks Doe you doubt that?
    
    \laerspeaks For \ham, and the trifling of his fauours,\\
    Hold it a fashion and a toy in Bloud;\\
    A Violet in the youth of Primy Nature;\\
    Froward, not permanent; sweet not lasting\\
    The suppliance of a minute? No more.
    
    \ophespeaks No more but so.
    
    \laerspeaks Thinke it no more:\\
    For nature cressant does not grow alone,\\
    In thewes and Bulke: but as his Temple waxes,\\
    The inward seruice of the Minde and Soule\\
    Growes wide withall. Perhaps he loues you now,\\
    And now no soyle nor cautell doth besmerch\\
    The vertue of his feare: but you must feare
    
    His greatnesse weigh'd, his will is not his owne;\\
    For hee himselfe is subiect to his Birth:\\
    Hee may not, as vnuallued persons doe,\\
    Carue for himselfe; for, on his choyce depends\\
    The sanctity and health of the weole State.\\
    And therefore must his choyce be circumscrib'd\\
    Vnto the voyce and yeelding of that Body,\\
    Whereof he is the Head. Then if he sayes he loues you,\\
    It fits your wisedome so farre to beleeue it;\\
    As he in his peculiar Sect and force\\
    May giue his saying deed: which is no further,\\
    Then the maine voyce of Denmarke goes withall.\\
    Then weigh what losse your Honour may sustaine,\\
    If with too credent eare you list his Songs;\\
    Or lose your Heart; or your chast Treasure open\\
    To his vnmastred importunity.\\
    Feare it \ophe, feare it my deare Sister,\\
    And keepe within the reare of your Affection;\\
    Out of the shot and danger of Desire.\\
    The chariest Maid is Prodigall enough,\\
    If she vnmaske her beauty to the Moone:\\
    Vertue it selfe scapes not calumnious stroakes,\\
    The Canker Galls, the Infants of the Spring\\
    Too oft before the buttons be disclos'd,\\
    And in the Morne and liquid dew of Youth,\\
    Contagious blastments are most imminent.\\
    Be wary then, best safety lies in feare;\\
    Youth to it selfe rebels, though none else neere.\\
    Ophe. I shall th'effect of this good Lesson keepe,\\
    As watchmen to my heart: but good my Brother\\
    Doe not as some vngracious Pastors doe,\\
    Shew me the steepe and thorny way to Heauen;\\
    Whilst like a puft and recklesse Libertine\\
    Himselfe, the Primrose path of dalliance treads,\\
    And reaks not his owne reade.
    
    \laerspeaks Oh, feare me not.\\
    (\textit{Enter Polonius.})\\
    I stay too long; but here my Father comes:
    
    A double blessing is a double grace;\\
    Occasion smiles vpon a second leaue.
    
    \polspeaks Yet heere \laer? Aboord, aboord for shame,\\
    The winde sits in the shoulder of your saile,\\
    And you are staid for there: my blessing with you;\\
    And these few Precepts in thy memory,\\
    See thou Character. Giue thy thoughts no tongue,\\
    Nor any vnproportion'd thought his Act:\\
    Be thou familiar; but by no meanes vulgar:\\
    The friends thou hast, and their adoption tride,\\
    Grapple them to thy Soule, with hoopes of Steele:\\
    But doe not dull thy palme, with entertainment\\
    Of each vnhatch't, vnfledg'd Comrade. Beware\\
    Of entrance to a quarrell: but being in\\
    Bear't that th'opposed may beware of thee.\\
    Giue euery man thine eare; but few thy voyce:\\
    Take each mans censure; but reserue thy iudgement:\\
    Costly thy habit as thy purse can buy;\\
    But not exprest in fancie; rich, not gawdie:\\
    For the Apparell oft proclaimes the man.\\
    And they in France of the best ranck and station,\\
    Are of a most select and generous cheff in that.\\
    Neither a borrower, nor a lender be;\\
    For lone oft loses both it selfe and friend:\\
    And borrowing duls the edge of Husbandry.\\
    This aboue all; to thine owne selfe be true:\\
    And it must follow, as the Night the Day,\\
    Thou canst not then be false to any man.\\
    Farewell: my Blessing season this in thee.
    
    \laerspeaks Most humbly doe I take my leaue, my Lord.
    
    \polspeaks The time inuites you, goe, your seruants tend.
    
    \laerspeaks Farewell \ophe, and remember well\\
    What I haue said to you.
    
    \ophespeaks Tis in my memory lockt,\\
    And you your selfe shall keepe the key of it.
    
    \laerspeaks Farewell. \hfill(\textit{Exit \laer.}
    
    \polspeaks What ist \ophe he hath said to you?
    
    \ophespeaks So please you, somthing touching the L. \ham.
    
    \polspeaks Marry, well bethought:\\
    Tis told me he hath very oft of late\\
    Giuen priuate time to you; and you your selfe\\
    Haue of your audience beene most free and bounteous.\\
    If it be so, as so tis put on me;\\
    And that in way of caution: I must tell you,\\
    You doe not vnderstand your selfe so cleerely,\\
    As it behoues my Daughter, and your Honour.\\
    What is betweene you, giue me vp the truth?
    
    \ophespeaks He hath my Lord of late, made many tenders\\
    Of his affection to me.
    
    \polspeaks Affection, puh. You speake like a greene Girle,\\
    Vnsifted in such perillous Circumstance.\\
    Doe you beleeue his tenders, as you call them?
    
    \ophespeaks I do not know, my Lord, what I should thinke.
    
    \polspeaks Marry Ile teach you; thinke your selfe a Baby,\\
    That you haue tane his tenders for true pay,\\
    Which are not starling. Tender your selfe more dearly;\\
    Or not to crack the winde of the poore Phrase,\\
    Roaming it thus, you'l tender me a foole.
    
    \ophespeaks My Lord, he hath importun'd me with loue,\\
    In honourable fashion.
    
    \polspeaks I, fashion you may call it, go too, go too.
    
    \ophespeaks And hath giuen countenance to his speech,\\
    My Lord, with all the vowes of Heauen.
    
    \polspeaks I, Springes to catch Woodcocks. I doe know\\
    When the Bloud burnes, how Prodigall the Soule\\
    Giues the tongue vowes: these blazes, Daughter,\\
    Giuing more light then heate; extinct in both,\\
    Euen in their promise, as it is a making;\\
    You must not take for fire. For this time Daughter,\\
    Be somewhat scanter of your Maiden presence;\\
    Set your entreatments at a higher rate,\\
    Then a command to parley. For Lord \ham,\\
    Beleeue so much in him, that he is young,\\
    And with a larger tether may he walke,\\
    Then may be giuen you. In few, \ophe,\\
    Doe not beleeue his vowes; for they are Broakers,\\
    Not of the eye, which their Inuestments show:\\
    But meere implorators of vnholy Sutes,\\
    Breathing like sanctified and pious bonds,\\
    The better to beguile. This is for all:\\
    I would not, in plaine tearmes, from this time forth,\\
    Haue you so slander any moment leisure,\\
    As to giue words or talke with the Lord \ham:\\
    Looke too't, I charge you; come your wayes.
    
    \ophespeaks I shall obey my Lord. \hfill(\textit{Exeunt.})
\end{drama*}

\scene

\updatemarks

\StageDir{Enter \ham, \hor, \mar.}

\begin{drama*}
    \hamspeaks The Ayre bites shrewdly: is it very cold?
    
    \horspeaks It is a nipping and an eager ayre.
    
    \hamspeaks What hower now?
    
    \horspeaks I thinke it lacks of twelue.
    
    \marspeaks No, it is strooke.
    
    \horspeaks Indeed I heard it not: then it drawes neere the season,\\
    Wherein the Spirit held his wont to walke.
    
    What does this meane my Lord?
    
    \hamspeaks The King doth wake to night, and takes his rouse,\\
    Keepes wassels and the swaggering vpspring reeles,\\
    And as he dreines his draughts of Renish downe,\\
    The kettle Drum and Trumpet thus bray out\\
    The triumph of his Pledge.
    
    \horspeaks Is it a custome?
    
    \hamspeaks I marry ist;\\
    And to my mind, though I am natiue heere,\\
    And to the manner borne: It is a Custome\\
    More honour'd in the breach, then the obseruance.
    
    \direct{Enter Ghost.}
    
    \horspeaks Looke my Lord, it comes.
    
    \hamspeaks Angels and Ministers of Grace defend vs:\\
    Be thou a Spirit of health, or Goblin damn'd,\\
    Bring with thee ayres from Heauen, or blasts from Hell,\\
    Be thy euents wicked or charitable,\\
    Thou com'st in such a questionable shape\\
    That I will speake to thee. Ile call thee \ham,\\
    King, Father, Royall Dane: Oh, oh, answer me,\\
    Let me not burst in Ignorance; but tell\\
    Why thy Canoniz'd bones Hearsed in death,\\
    Haue burst their cerments, why the Sepulcher\\
    Wherein we saw thee quietly enurn'd,\\
    Hath op'd his ponderous and Marble iawes,\\
    To cast thee vp againe? What may this meane?\\
    That thou dead Coarse againe in compleat steele,\\
    Reuisits thus the glimpses of the Moone,\\
    Making Night hidious? And we fooles of Nature,\\
    So horridly to shake our disposition,\\
    With thoughts beyond thee; reaches of our Soules,\\
    Say, why is this? wherefore? what should we doe?\\
    (\textit{Ghost beckens \ham.})
    
    \horspeaks It beckons you to goe away with it,\\
    As if it some impartment did desire\\
    To you alone.
    
    \marspeaks Looke with what courteous action\\
    It wafts you to a more remoued ground:\\
    But doe not goe with it.
    
    \horspeaks No, by no meanes.
    
    \hamspeaks It will not speake: then will I follow it.
    
    \horspeaks Doe not my Lord.
    
    \hamspeaks Why, what should be the feare?\\
    I doe not set my life at a pins fee;\\
    And for my Soule, what can it doe to that?\\
    Being a thing immortall as it selfe:\\
    It waues me forth againe; Ile follow it.
    
    \horspeaks What if it tempt you toward the Floud my Lord?\\
    Or to the dreadfull Sonnet of the Cliffe,\\
    That beetles o're his base into the Sea,\\
    And there assumes some other horrible forme,\\
    Which might depriue your Soueraignty of Reason,\\
    And draw you into madnesse thinke of it?
    
    \hamspeaks It wafts me still: goe on, Ile follow thee.
    
    \marspeaks You shall not goe my Lord.
    
    \hamspeaks Hold off your hand.
    
    \horspeaks Be rul'd, you shall not goe.
    
    \hamspeaks My fate cries out,\\
    And makes each petty Artire in this body,\\
    As hardy as the Nemian Lions nerue:\\
    Still am I cal'd? Vnhand me Gentlemen:\\
    By Heau'n, Ile make a Ghost of him that lets me:\\
    I say away, goe on, Ile follow thee.
    
    \direct{Exeunt \gho \& \ham.}
    
    \horspeaks He waxes desperate with imagination.
    
    \marspeaks Let's follow; 'tis not fit thus to obey him.
    
    \horspeaks Haue after, to what issue will this come?
    
    \marspeaks Something is rotten in the State of Denmarke.
    
    \horspeaks Heauen will direct it.
    
    \marspeaks Nay, let's follow him. \hfill(\textit{Exeunt.})
\end{drama*}

\scene

\updatemarks

\StageDir{Enter \gho and \ham.}

\begin{drama*}
    \hamspeaks Where wilt thou lead me? speak; Ile go no further.
    
    \ghospeaks Marke me.
    
    \hamspeaks I will.
    
    \ghospeaks My hower is almost come,\\
    When I to sulphurous and tormenting Flames\\
    Must render vp my selfe.
    
    \hamspeaks Alas poore Ghost.
    
    \ghospeaks Pitty me not, but lend thy serious hearing\\
    To what I shall vnfold.
    
    \hamspeaks Speake, I am bound to heare.
    
    \ghospeaks So art thou to reuenge, when thou shalt heare.
    
    \hamspeaks What?
    
    \ghospeaks I am thy Fathers Spirit,\\
    Doom'd for a certaine terme to walke the night;\\
    And for the day confin'd to fast in Fiers,\\
    Till the foule crimes done in my dayes of Nature\\
    Are burnt and purg'd away? But that I am forbid\\
    To tell the secrets of my Prison-House;\\
    I could a Tale vnfold, whose lightest word\\
    Would harrow vp thy soule, freeze thy young blood,\\
    Make thy two eyes like Starres, start from their Spheres,\\
    Thy knotty and combined lockes to part,\\
    And each particular haire to stand an end,\\
    Like Quilles vpon the fretfull Porpentine:\\
    But this eternall blason must not be\\
    To eares of flesh and bloud; list \ham, oh list,\\
    If thou didst euer thy deare Father loue.
    
    \hamspeaks Oh Heauen!
    
    \ghospeaks Reuenge his foule and most vnnaturall Murther.
    
    \hamspeaks Murther?
    
    \ghospeaks Murther most foule, as in the best it is;\\
    But this most foule, strange, and vnnaturall.
    
    \hamspeaks Hast, hast me to know it,\\
    That with wings as swift\\
    As meditation, or the thoughts of Loue,\\
    May sweepe to my Reuenge.
    
    \ghospeaks I finde thee apt,\\
    And duller should'st thou be then the fat weede\\
    That rots it selfe in ease, on Lethe Wharfe,\\
    Would'st thou not stirre in this. Now \ham heare:\\
    It's giuen out, that sleeping in mine Orchard,\\
    A Serpent stung me: so the whole eare of Denmarke,\\
    Is by a forged processe of my death\\
    Rankly abus'd: But know thou Noble youth,\\
    The Serpent that did sting thy Fathers life,\\
    Now weares his Crowne.
    
    \hamspeaks O my Propheticke soule: mine Vncle?
    
    \ghospeaks I that incestuous, that adulterate Beast\\
    With witchcraft of his wits, hath Traitorous guifts.\\
    Oh wicked Wit, and Gifts, that haue the power\\
    So to seduce? Won to to this shamefull Lust\\
    The will of my most seeming vertuous Queene:\\
    Oh \ham, what a falling off was there,\\
    From me, whose loue was of that dignity,\\
    That it went hand in hand, euen with the Vow\\
    I made to her in Marriage; and to decline\\
    Vpon a wretch, whose Naturall gifts were poore\\
    To those of mine. But Vertue, as it neuer wil be moued,\\
    Though Lewdnesse court it in a shape of Heauen:\\
    So Lust, though to a radiant Angell link'd,\\
    Will sate it selfe in a Celestiallbed, \& prey on Garbage.
    
    But soft, me thinkes I sent the Mornings Ayre;\\
    Briefe let me be: Sleeping within mine Orchard,\\
    My custome alwayes in the afternoone;\\
    Vpon my secure hower thy Vncle stole\\
    With iuyce of cursed Hebenon in a Violl,\\
    And in the Porches of mine eares did poure\\
    The leaperous Distilment; whose effect\\
    Holds such an enmity with bloud of Man,\\
    That swift as Quick-siluer, it courses through\\
    The naturall Gates and Allies of the Body;\\
    And with a sodaine vigour it doth posset\\
    And curd, like Aygre droppings into Milke,\\
    The thin and wholsome blood: so did it mine;\\
    And a most instant Tetter bak'd about,\\
    Most Lazar-like, with vile and loathsome crust,\\
    All my smooth Body.\\
    Thus was I, sleeping, by a Brothers hand,\\
    Of Life, of Crowne, and Queene at once dispatcht;\\
    Cut off euen in the Blossomes of my Sinne,\\
    Vnhouzzled, disappointed, vnnaneld,\\
    No reckoning made, but sent to my account\\
    With all my imperfections on my head;\\
    Oh horrible, Oh horrible, most horrible:\\
    If thou hast nature in thee beare it not;\\
    Let not the Royall Bed of Denmarke be\\
    A Couch for Luxury and damned Incest.\\
    But howsoeuer thou pursuest this Act,\\
    Taint not thy mind; nor let thy Soule contriue\\
    Against thy Mother ought; leaue her to heauen,\\
    And to those Thornes that in her bosome lodge,\\
    To pricke and sting her. Fare thee well at once;\\
    The Glow-worme showes the Matine to be neere.\\
    And gins to pale his vneffectuall Fire:\\
    Adue, adue, \ham: remember me. \hfill(\textit{Exit.})
    
    \hamspeaks Oh all you host of Heauen! Oh Earth; what els?\\
    And shall I couple Hell? Oh fie: hold my heart;\\
    And you my sinnewes, grow not instant Old;\\
    But beare me stiffely vp: Remember thee?\\
    I, thou poore Ghost, while memory holds a seate\\
    In this distracted Globe: Remember thee?\\
    Yea, from the Table of my Memory,\\
    Ile wipe away all triuiall fond Records,\\
    All sawes of Bookes, all formes, all presures past.\\
    That youth and obseruation coppied there;\\
    And thy Commandment all alone shall liue\\
    Within the Booke and Volume of my Braine,\\
    Vnmixt with baser matter; yes, yes, by Heauen:\\
    Oh most pernicious woman!\\
    Oh Villaine, Villaine, smiling damned Villaine!\\
    My Tables, my Tablet; meet it is I set it downe,\\
    That one may smile, and smile and be a Villaine!\\
    At least I'm sure it may be so in Denmarke;\\
    So Vnckle there you are: now to my word;\\
    It is; Adue, Adue, Remember me: I haue sworn't.\\
    \mbox{}\\
    (\textit{\hor \& \mar within.})\\
    My Lord, my Lord,
    
    \direct{Enter \hor And \mar}
    
    \marspeaks Lord \ham.
    
    \horspeaks Heauen secure him.
    
    \marspeaks So be it.'
    
    \horspeaks Illo, ho, ho, my Lord.'
    
    \hamspeaks Hillo, ho, ho, boy; come bird, come.
    
    \marspeaks How ist't my Noble Lord?
    
    \horspeaks What newes, my Lord?
    
    \hamspeaks Oh wonderful!
    
    \horspeaks Good my Lord tell it.
    
    \hamspeaks No you'l reueale it.
    
    \horspeaks Not I, my Lord, by Heauen.
    
    \marspeaks Nor I, my Lord.
    
    \hamspeaks How say you then, would heart of man once think it?\\
    But you'l be secret?
    
    \speaker{\mar and \barn} I, by Heau'n, my Lord.
    
    \hamspeaks There's nere a villaine dwelling in all Denmarke\\
    But hee's an arrant knaue.
    
    \horspeaks There needs no Ghost my Lord, come from the\\
    Graue, to tell vs this.
    
    \hamspeaks Why right, you are i'th'right;\\
    And so, without more circumstance at all,\\
    I hold it fit that we shake hands and part;\\
    You,as your busines and desires shall point you:\\
    For euery man ha's businesse and desire,\\
    Such as it is; and for mine owne poore part,\\
    Looke you, Ile goe pray.
    
    \horspeaks These are but wild and hurling words, my Lord.
    
    \hamspeaks I'm sorry they offend you heartily:\\
    Yes faith, heartily.
    
    \horspeaks There's no offence my Lord.
    
    \hamspeaks Yes, by Saint Patricke, but there is my Lord,\\
    And much offence too, touching this Vision heere:\\
    It is an honest Ghost, that let me tell you:\\
    For your desire to know what is betweene vs,\\
    O'remaster't as you may. And now good friends,\\
    As you are Friends, Schollers and Soldiers.\\
    Giue me one poore request.
    
    \horspeaks What is't my Lord? we will.
    
    \hamspeaks Neuer make known what you haue seen to night.
    
    \speaker{\mar and \barn} My Lord, we will not.
    
    \hamspeaks Nay, but swear't.
    
    \horspeaks In faith my Lord, not I.
    
    \marspeaks Nor I my Lord: in faith.
    
    \hamspeaks Vpon my sword.
    
    \marspeaks We haue sworne my Lord already.
    
    \hamspeaks Indeed, vpon my sword Indeed.
    
    \ghospeaks Sweare. \hfill(\textit{\gho cries vnder the Stage.})
    
    \hamspeaks Ah ha boy, sayest thou so. Art thou there true-penny?\\
    Come one you here this fellow in the selleredge\\
    Consent to sweare.
    
    \horspeaks Propose the Oath my Lord.
    
    \hamspeaks Neuer to speake of this that you haue seene.\\
    Sweare by my sword.
    
    \ghospeaks Sweare.
    
    \hamspeaks Hic \& vbique? Then wee'l shift for grownd.\\
    Come hither Gentlemen,\\
    And lay your hands againe vpon my sword,\\
    Neuer to speake of this that you haue heard:\\
    Sweareby my Sword.
    
    \ghospeaks Sweare.
    
    \hamspeaks Well said old Mole, can'st worke i'th'ground so fast?\\
    A worthy Pioner, once more remoue good friends.
    
    \horspeaks Oh day and night: but this is wondrous strange.
    
    \hamspeaks And therefore as a stranger giue it welcome.\\
    There are more things in Heauen and Earth, Horatio,\\
    Then are dream't of in our Philosophy. But come,\\
    Here as before, neuer so helpe you mercy,\\
    How strange or odde so ere I beare my selfe;\\
    (As I perchance heereafter shall thinke meet\\
    To put an Anticke disposition on :)\\
    That you at such time seeing me, neuer shall\\
    With Armes encombred thus, or thus, head shake;\\
    Or by pronouncing of some doubtful! Phrase;\\
    As well, we know, or we could and if we would,\\
    Or if we lift to speake; or there be and if there might,\\
    Or such ambiguous giuing out to note,
    
    That you know ought of me; this not to doe:\\
    So grace and mercy at your most neede helpe you\\
    Sweare.
    
    \ghospeaks Sweare.
    
    \hamspeaks Rest, Rest perturbed Spirit: so Gentlemen,\\
    With all my loue I doe commend me to you;\\
    And what so poore a man as \ham is,\\
    May doe t'expresse his loue and friending to you,\\
    God willing shall not lacke: let vs goe in together,\\
    And still your fingers on your lippes I pray,\\
    The time is out of ioynt: Oh cursed spight,\\
    That euer I was borne to set it right,\\
    Nay, come let's goe together. \hfill(\textit{Exeunt.})
\end{drama*}

\act

\scene

\updatemarks

\StageDir{Enter \pol and \reyn.}

\begin{drama*}
    \polspeaks Giue him his money,and these notes \reyn.
    
    \reynspeaks I will my Lord.
    
    \polspeaks You shall doe maruels wisely: good \reyn\\
    Before you visite him you make inquiry\\
    Of his behauiour.
    
    \reynspeaks My Lord, I did intend it.
    
    \polspeaks Matty, well said;\\
    Very well said. Looke you Sir,\\
    Enquire me first what Danskers are in Paris;\\
    And how, and who; what meanes; and where they keepe:\\
    What company, at what expence: and finding\\
    By this encompassement and drift of question\\
    That they doe know my sonne: Come you more neerer\\
    Then your particular demands will touch it,\\
    Take you as 'twere some distant knowledge of him.\\
    And thus I know his father and his friends,\\
    And in part him. Doe you marke this \reyn?
    
    \reynspeaks I, very well my Lord.
    
    % typo in wikisource (Pelon.)
    \polspeaks And in part him, but you may say not well;\\
    But if't be hee I meane, hees very wilde;\\
    Addicted so and so; and there put on him\\
    What forgeries you please: marry, none so ranke,\\
    As may dishonour him; take heed of that:\\
    But Sir, such wanton, wild, and vsuall flips,\\
    As are Companions noted and most knowne\\
    To youth and liberty.
    
    \reynspeaks As gaming my Lord.
    
    % typo in wikisource (Poton)
    \polspeaks I, or drinking, fencing, swearing,\\
    Quarelling, drabbing. You may goe so farre.
    
    \reynspeaks My Lord that would dishonour him.
    
    \polspeaks Faith no, as you may season it in the charge;\\
    You must not put another scandall on him,\\
    That hee is open to lncontinencie;\\
    That's not my meaning; but breath his faults so quaintly.\\
    That they may seeme the taints of liberty.\\
    The flash and out-breake of a fiery minde,\\
    A sauagenes in vnreclaim'd bloud of generall affault.
    
    \reynspeaks But my good Lord.
    
    \polspeaks Wherefore should you doe this?
    
    \reynspeaks I my Lord, I would know that.
    
    \polspeaks Many Sir, heere's my drift,\\
    And I belieue it is a fetch of warrant:\\
    Yon laying these flight full eyes on my Sonne,\\
    As 'twere a thing a little soil'd i'th'working:\\
    Marke you your party in conuerse; him you would sound\\
    Hauing euer seene. In the prenominate crimes,\\
    The youth you breath of guilty, be assur'd\\
    He closes with you in this consequence:\\
    Good sir, or so, or friend, or Gentleman.\\
    According to the Phrase and the Addition,\\
    Of man and Country.
    
    \reynspeaks Very good my Lord.
    
    \polspeaks And then Sir does he this?\\
    He does: what was I about to say?\\
    I was about to say somthing: where did I leaue?
    
    \reynspeaks At closes in the consequence:\\
    At friend, or so, and Gentleman.
    
    \polspeaks At closes in the consequence, I marry.\\
    He closes with you thus. I know the Gentleman,\\
    I saw him yesterday, or tother day;\\
    Or then or then, with such and such; and as you say.\\
    There was he gaming, there o'retooke in's Rouse,\\
    There falling out at Tennis; or perchance,\\
    I saw him enter such a house of saile;\\
    Videlicet, a Brothell, or so forth. See you now;\\
    Your bait of falshood, takes this Cape of truth;\\
    And thus doe we of wisedome and of reach\\
    With windlesses, and with assaies of Bias,\\
    By indirections finde directions out:\\
    So by my former Lecture and aduice\\
    Shall you my Sonne; you haue me, haue you not?
    
    \reynspeaks My Lord I haue.
    
    \polspeaks God buy you; fare you well
    
    \reynspeaks Good my Lord.
    
    \polspeaks Obserue his inclination in your selfe.
    
    \reynspeaks I shall my Lord.
    
    \polspeaks And let him plye his Musicke.
    
    \reynspeaks Well, my Lord. \hfill(\textit{Exit.})
    
    \direct{Enter \ophe.}
    
    \polspeaks Farewell:\\
    How now \ophe, what's the matter?
    
    \ophespeaks Alas my Lord,I haue beene so affrighted.
    
    \polspeaks With what, in the name of Heauen?
    
    \ophespeaks My Lord, as I was sowing in my Chamber,\\
    Lord Hamlet with his doublet all vnbrac'd,\\
    No hat vpon his head, his stockings foul'd,\\
    Vngartred, and downe giued to his Anckle,\\
    Pale as bis shirt, his knees knocking each other,\\
    And with a looke so pitious in purport,\\
    As if he had been loosed out of hell,\\
    To speake of horrors: he comes before me.
    
    \polspeaks Mad for thy Loue?
    
    \ophespeaks My Lord, I doe not know; but truly I do feare it.
    
    \polspeaks What said he?
    
    \ophespeaks He tooke me by the wrist, and held me hard;\\
    Then goes he to the length of all his arme;\\
    And with his other hand thus o're his brow,\\
    He fals to such perusall of my face,\\
    As he would draw it. Long staid he so,\\
    At last, a little shaking of mine Arme:\\
    And thrice his head thus wauing vp and downe,\\
    He rais'd a sigh, so pittious and profound,\\
    That it did seeme to shatter all his bulke.\\
    And end his being. That done, he lets me goe.\\
    And with his head ouer his shoulders turn'd,\\
    He seem'd to finde his way without his eyes.\\
    For out adores he went without their helpe;\\
    And to the last, bended their light on me.
    
    \polspeaks Goe with me. I will goe seeke the King,\\
    This is the very extasie of Loue,\\
    Whose violent property foredoes it selfe,\\
    And leads the will to desperate Vndertakings,\\
    As oft as any passion vnder Heauen,\\
    That does afflict our Natures. I am sorrie,\\
    What haue you giuen him any hard words of late?
    
    \ophespeaks No my good Lord: but as you did command,\\
    I did repell his Letters, and deny'de\\
    His accesse to me.
    
    \polspeaks That hath made him mad.\\
    I am sorrie that with better speed and iudgement\\
    I had not quoted him. I feare he did but trifle,\\
    And meant to wracke thee: but beshrew my iealousie:\\
    It seemes it is as proper to our Age,\\
    To cast beyond our selues in our Opinions,\\
    As it is common for the yonger sort\\
    To lacke discretion. Come, go we to the King,\\
    This must be knowne, wc being kept close might moue\\
    More greefe to hide, then hate to vtter loue. \hfill(\textit{Exeunt.})
\end{drama*}

\scene

\updatemarks

\StageDir{Enter \king, \queen, \rosin, and \guil Cumalijs.}

\begin{drama*}
    \kingspeaks Welcome deere \rosin and \guil.\\
    Moreouer, that we much did long to see you,\\
    The neede we haue to vse you, did prouoke\\
    Our hastie sending. Something haue you heard\\
    Of Hamlets transformation: so I call it,\\
    Since not th'exterior, nor the inward man\\
    Resembles that it was. What it should bee\\
    More then his Fathers death, that thus hath put him\\
    So much from th'vnderstanding of himselfe,\\
    I cannot deeme of. I intreat you both,\\
    That being of so young dayes brought vp with him:\\
    And since so Neighbour'd to his youth, and humour,\\
    That you vouchsafe your rest heere in our Court\\
    Some little time: so by your Companies\\
    To draw him on to pleasures, and to gather\\
    So much as from Occasions you may gleane,\\
    That open'd lies within our remedie.
    
    \queenspeaks Good Gentlemen, he hath much talk'd of you,\\
    And sure I am, two men there are not liuing,\\
    To whom he more adheres. If it will please you\\
    To shew vs so much Gentrie, and good will,\\
    As to expend your time with vs a‑while,\\
    For the supply and profit of our Hope,\\
    Your Visitation shall receiue such thankes\\
    As fits a Kings remembrance.
    
    \rosinspeaks Both your Maiesties\\
    Might by the Soueraigne power you haue of vs,\\
    Put your dread pleasures, more into Command\\
    Then to Entreatie.
    
    \guilspeaks We both obey,\\
    And here giue vp our selues, in the full bent,\\
    To lay our Seruices freely at your feete,\\
    To be commanded.
    
    \kingspeaks Thankes \rosin, and gentle \guil.
    
    \queenspeaks Thankes \guil and gentle \rosin.\\
    And I beseech you instantly to visit\\
    My too much changed Sonne.\\
    Go some of ye,\\
    And bring the Gentlemen where \ham is.
    
    \guilspeaks Heauens make our presence and our practices\\
    Pleasant and helpfull to him. \hfill(\textit{Exit.})
    
    \queenspeaks Amen.
    
    \direct{Enter \pol.}
    
    \polspeaks Th'Ambassadors from Norwey, my good Lord,\\
    Are ioyfully return'd.
    
    \kingspeaks Thou still hast bin the Father of good Newes.
    
    \polspeaks Haue I, my Lord? Assure you, my good Liege,\\
    I hold my dutie, as I hold my Soule,\\
    Both to my God, one to my gracious King:\\
    And I do thinke, or else this braine of mine\\
    Hunts not the traile of Policie, so sure\\
    As I haue vs'd to do: that I haue found\\
    The very cause of Hamlets Lunacie.
    
    \kingspeaks Oh speake of that, that I do long to heare.
    
    \polspeaks Giue first admittance to th'Ambassadors,\\
    My Newes shall be the Newes to that great Feast.
    
    \kingspeaks Thy selfe do grace to them, and bring them in.\\
    He tels me my sweet Queene, that he hath found\\
    The head and sourse of all your Sonnes distemper.
    
    \queenspeaks I doubt it is no other, but the maine,\\
    His Fathers death, and our o're‑hasty Marriage.
    
    \direct{Enter \pol, \textsc{Uoltumand}, and \corn.}
    
    \kingspeaks Well, we shall sift him. Welcome good Frends:\\
    Say \volt, what from our Brother Norwey?
    
    \voltspeaks Most faire returne of Greetings, and Desires.\\
    Vpon our first, he sent out to suppresse\\
    His Nephewes Leuies, which to him appear'd\\
    To be a preparation 'gainst the Poleak:\\
    But better look'd into, he truly found\\
    It was against your Highnesse, whereat greeued,\\
    That so his Sicknesse, Age, and Impotence\\
    Was falsely borne in hand, sends out Arrests\\
    On Fortinbras, which he (in breefe) obeyes,\\
    Receiues rebuke from Norwey: and in fine,\\
    Makes Vow before his Vnkle, neuer more\\
    To giue th'assay of Armes against your Maiestie.\\
    Whereon old Norwey, ouercome with ioy,\\
    Giues him three thousand Crownes in Annuall Fee,\\
    And his Commission to imploy those Soldiers\\
    So leuied as before, against the Poleak:\\
    With an intreaty heerein further shewne,\\
    That it might please you to giue quiet passe\\
    Through your Dominions, for his Enterprize,\\
    On such regards of safety and allowance,\\
    As therein are set downe.
    
    \kingspeaks It likes vs well:\\
    And at our more consider'd time wee'l read,\\
    Answer, and thinke vpon this Businesse.\\
    Meane time we thanke you, for your well‑tooke Labour.\\
    Go to your rest, at night wee'l Feast together.\\
    Most welcome home. \hfill(\textit{Exit Ambass.})
    
    \polspeaks This businesse is very well ended.\\
    My Liege, and Madam, to expostulate\\
    What Maiestie should be, what Dutie is,\\
    Why day is day; night, night; and time is time,\\
    Were nothing but to waste Night, Day and Time.\\
    Therefore, since Breuitie is the Soule of Wit,\\
    And tediousnesse, the limbes and outward flourishes,\\
    I will be breefe. Your Noble Sonne is mad:\\
    Mad call I it; for to define true Madnesse,\\
    What is't, but to be nothing else but mad.\\
    But let that go.
    
    \queenspeaks More matter, with lesse Art.
    
    \polspeaks Madam I sweare I vse no Art at all:\\
    That he is mad, 'tis true: 'Tis true 'tis pittie,\\
    And pittie it is true: A foolish figure,\\
    But farewell it: for I will vse no Art.\\
    Mad let vs grant him then: and now remaines\\
    That we finde out the cause of this effect,\\
    Or rather say, the cause of this defect;\\
    For this effect defectiue, comes by cause,\\
    Thus it remaines, and the remainder thus. Perpend,\\
    I haue a daughter: haue, whil'st she is mine,\\
    Who in her Dutie and Obedience, marke,\\
    Hath giuen me this: now gather, and surmise.\\
    \mbox{}\\
    (\textit{The Letter.})\\
    \textit{To the Celestiall, and my Soules Idoll, the most beautified \ophe.}\\
    That's an ill Phrase, a vilde Prase, beautified is a vilde\\
    Phrase: but you shall heare these in her excellent white bosome, these.
    
    \queenspeaks Came this from \ham to her.
    
    \polspeaks \textit{Good Madam stay awhile, I will be faithfull.\\
    Doubt thou, the Starres are fire,\\
    Doubt, that the Sunne doth moue:\\
    Doubt Truth to be a Lier,\\
    But nener Doubt, I loue.\\
    O deere \ophe, I am ill at these Numbers: I haue not Art to\\
    reckon my grones; but that I loue thee best, oh most Best be\\
    leeue it. Adieu.}\\\vspace{6pt}
    \hspace{8em}\parbox{20em}{\textit{Thine euermore most deere Lady, whilst this\\
    Machine is to him, \ham.}}\\
    \mbox{}\\
    This in Obedience hath my daughter shew'd me:\\
    And more aboue hath his soliciting,\\
    As they fell out by Time, by Meanes, and Place,\\
    All giuen to mine eare.
    
    \kingspeaks But how hath she receiu'd his Loue?
    
    \polspeaks What do you thinke of me?
    
    \kingspeaks As of a man, faithfull and Honourable.
    
    \polspeaks I wold faine proue so. But what might you think?\\
    When I had seene this hot loue on the wing,\\
    As I perceiued it, I must tell you that\\
    Before my Daughter told me, what might you\\
    Or my [deere] Maiestie your Queene heere, think,\\
    If I had playd the Deske or Table-booke,\\
    Or giuen my heart a winking, mute and dumbe,\\
    Or look'd vpon this Loue, with idle sight,\\
    What might you thinke? No, I went round to worke,\\
    And (my yong Mistris) thus I did bespeake\\
    Lord Hamlet is, a Prince out of thy Starre,\\
    This must not be: and then, I Precepts gaue her,\\
    That she should locke her selfe from his Resort,\\
    Admit no Messengers, receiue no Tokens:\\
    Which done, she tooke the Fruites of my Aduice,\\
    And he repulsed. A short Tale to make,\\
    Fell into a Sadnesse, then into a Fast,\\
    Thence to a Watch, thence into a Weaknesse,\\
    Thence to a Lightnesse, and by this declension\\
    Into the Madnesse whereon now he raues,\\
    And all we waile for.
    
    \kingspeaks Do you thinke 'tis this?
    
    \queenspeaks It may be very likely.
    
    \polspeaks Hath there bene such a time, I'de fain know that,\\
    That I haue possitiuely said, 'tis so,\\
    When it prou'd otherwise?
    
    \kingspeaks Not that I know.
    
    \polspeaks Take this from this; if this be otherwise,\\
    If Circumstances leade me, I will finde\\
    Where truth is hid, though it were hid indeede\\
    Within the Center.
    
    \kingspeaks How may we try it further?
    
    \polspeaks You know sometimes\\
    He walkes foure houres together, here\\
    In the Lobby.
    
    \queenspeaks So he ha's indeed.
    
    \polspeaks At such a time Ile loose my Daughter to him,\\
    Be you and I behinde an Arras then,\\
    Marke the encounter: If he loue her not,\\
    And be not from his reason falne thereon;\\
    Let me be no Assistant for a State,\\
    And keepe a Farme and Carters.
    
    \kingspeaks We will try it.
    
    \direct{Enter \ham reading on a Booke.}
    
    \queenspeaks But looke where sadly the poore wretch\\
    Comes reading.
    
    \polspeaks Away I do beseech you, both away,\\
    Ile boord him presently.\hfill(\textit{Exit \king \& \queen.}\\
    Oh giue me leaue. How does my good Lord \ham?
    
    \hamspeaks Well, God-a-mercy.
    
    \polspeaks Do you know me, my Lord?
    
    \hamspeaks Excellent, excellent well: y'are a Fishmonger.
    
    \polspeaks Not I my Lord.
    
    \hamspeaks Then I would you were so honest a man.
    
    \polspeaks Honest, my Lord?
    
    \hamspeaks I sir, to be honest as this world goes, is to bee one man pick'd out of two thousand.
    
    \polspeaks That's very true, my Lord.
    
    \hamspeaks For if the Sun breed Magots in a dead dogge, being a good kissing Carrion———\\
    Haue you a daughter?
    
    \polspeaks I haue my Lord.
    
    \hamspeaks Let her not walke i'th'Sunne; Conception is a blessing, but not as your daughter may conceiue. Friend looke too't.
    
    \polspeaks How say you by that? Still harping on my daughter: yet he knew me not at first; he said I was a Fishmonger: he is farre gone, farre gone: and truly in my youth, I suffred much extreamity for loue: very neere this. Ile speake to him againe. What do you read my Lord?
    
    \hamspeaks Words, words, words.
    
    \polspeaks What is the matter, my Lord?
    
    \hamspeaks Betweene who?
    
    \polspeaks I meane the matter you meane, my Lord.
    
    \hamspeaks Slanders Sir: for the Satyricall slaue saies here, that old men haue gray Beards; that their faces are wrinkled; their eyes purging thicke Amber, or Plum-Tree Gumme: and that they haue a plentifull locke of Wit, together with weake Hammes. All which Sir, though I most powerfully, and potently beleeue; yet I holde it not Honestie to haue it thus set downe: For you your selfe Sir, should be old as I am, if like a Crab you could go backward.
    
    \polspeaks Though this be madnesse,\\
    Yet there is Method in't: will you walke\\
    Out of the ayre my Lord?
    
    \hamspeaks Into my Graue?
    
    \polspeaks Indeed that is out o'th'Ayre:\\
    How pregnant (sometimes) his Replies are?\\
    A happinesse,\\
    That often Madnesse hits on,\\
    Which Reason and Sanitie could not\\
    So prosperously be deliuer'd of.\\
    I will leaue him,\\
    And sodainely contriue the meanes of meeting\\
    Betweene him, and my daughter.\\
    My Honourable Lord, I will most humbly\\
    Take my leaue of you
    
    \hamspeaks You cannot Sir take from me any thing, that I will more willingly part withall, except my life, my life.
    
    \polspeaks Fare you well my Lord.
    
    \hamspeaks These tedious old fooles.
    
    \polspeaks You goe to seeke my Lord \ham; there hee is.
    
    \direct{Enter \rosin and \guil.}
    
    \rosinspeaks God saue you Sir.
    
    \guilspeaks Mine honour'd Lord?
    
    \rosinspeaks My most deare Lord?
    
    \hamspeaks My excellent good friends? How do'st thou \guil? Oh, \rosin; good Lads: How doe ye both?
    
    \rosinspeaks As the indifferent Children of the earth.
    
    \guilspeaks Happy, in that we are not ouer-happy: on Fortunes Cap, we are not the very Button.
    
    \hamspeaks Nor the Soales of her Shoo?
    
    \rosinspeaks Neither my Lord.
    
    \hamspeaks Then you liue about her waste, or in the middle of her fauour?
    
    \guilspeaks Faith, her priuates, we.
    
    \hamspeaks In the secret parts of Fortune? Oh, most true:\\
    she is a Strumpet. What's the newes?
    
    \rosinspeaks None my Lord; but that the World's growne honest.
    
    \hamspeaks Then is Doomesday neere: But your newes is not true. Let me question more in particular: what haue you my good friends, deserued at the hands of Fortune, that she sends you to Prison hither?
    
    \guilspeaks Prison, my Lord?
    
    \hamspeaks Denmark's a Prison.
    
    \rosinspeaks Then is the World one.
    
    \hamspeaks A goodly one, in which there are many Confines, Wards and Dungeons; Denmarke being one o'th' worst.
    
    \rosinspeaks We thinke not so my Lord.
    
    \hamspeaks Why then 'tis none to you; for there is nothing either good or bad, but thinking makes it so: to me it is a prison.
    
    \rosinspeaks Why then your Ambition makes it one: 'tis too narrow for your minde.
    
    \hamspeaks O God, I could be bounded in a nutshell, and count my selfe a King of infinite space; were it not that I haue bad dreames.
    
    \guilspeaks Which dreames indeed are Ambition: for the very substance of the Ambitious, is meerely the shadow of a Dreame.
    
    \hamspeaks A dreame it selfe is but a shadow.
    
    \rosinspeaks Truely, and I hold Ambition of so ayry and light a quality, that it is but a shadowes shadow.
    
    \hamspeaks Then are our Beggers bodies; and our Monarchs and out-stretcht Heroes the Beggers Shadowes: shall wee to th'Court: for, by my fey I cannot reason?
    
    \speaker{\rosin and \guil} Wee'l wait vpon you.
    
    \hamspeaks No such matter. I will not sort you with the rest of my seruants: for to speake to you like an honest man: I am most dreadfully attended; but in the beaten way of friendship, What make you at Elsonower?
    
    \rosinspeaks To visit you my Lord, no other occasion.
    
    \hamspeaks Begger that I am, I am euen poore in thankes; but I thanke you: and sure deare friends my thanks are too deare a halfepeny; were you not sent for? Is it your owne inclining? Is it a free visitation? Come, deale iustly with me: come, come; nay speake.
    
    \guilspeaks What should we say my Lord?
    
    \hamspeaks Why any thing. But to the purpose; you were sent for; and there is a kinde confession in your lookes; which your modesties haue not craft enough to color, I know the good King \& Queene haue sent for you.
    
    \rosinspeaks To what end my Lord?
    
    \hamspeaks That you must teach me: but let mee coniure you by the rights of our fellowship, by the consonancy of our youth, by the Obligation of our euer-preserued loue, and by what more deare, a better proposer could charge you withall; be euen and direct with me, whether you were sent for or no.
    
    \rosinspeaks What say you?
    
    \hamspeaks Nay then I haue an eye of you; if you loue me hold not off.
    
    \guilspeaks My Lord, we were sent for.
    
    \hamspeaks I will tell you why; so shall my anticipation preuent your discouery of your secricie to the King and Queene: moult no feather, I haue of late, but wherefore I know not, lost all my mirth, forgone all custome of exercise; and indeed, it goes so heauenly with my disposition; that this goodly frame the Earth, seemes to me a sterrill Promontory; this most excellent Canopy the Ayre, look you, this braue ore-hanging, this Maiesticall Roofe, fretted with golden fire: why, it appeares no other thing to mee, then a foule and pestilent congregation of vapours. What a piece of worke is a man! how Noble in Reason? how infinite in faculty? in sorme and mouing how expresse and admirable? in Action, how like an Angel? in apprehension, how like a God? the beauty of the world, the Parragon of Animals; and yet to me, what is this Quintessence of Dust? Man delights not me; no, nor Woman neither; though by your smiling you seeme to say so.
    
    \rosinspeaks My Lord, there was no such stuffe in my thoughts.
    
    \hamspeaks Why did you laugh, when I said, Man delights not me?
    
    \rosinspeaks To thinke, my Lord, if you delight not in Man, what Lenton entertainment the Players shall receiue from you: wee coated them on the way, and hither are they comming to offer you Seruice.
    
    \hamspeaks He that playes the King shall be welcome; his Maiesty shall haue Tribute of mee: the aduenturous Knight shal vse his Foyle and Target: the Louer shall not sigh gratis, the humorous man shall end his part in peace: the Clowne shall make those laugh whose lungs are tickled a'th'sere: and the Lady shall say her minde freely; or the blanke Verse shall halt for't: what Players are they?
    
    \rosinspeaks Euen those you were wont to take delight in the Tragedians of the City.
    
    \hamspeaks How chances it they trauaile? their residence both in reputation and profit was better both wayes.
    
    \rosinspeaks I thinke their Inhibition comes by the meanes of the late Innouation?
    
    \hamspeaks Doe they hold the same estimation they did when I was in the City? Are they so follow'd?
    
    \rosinspeaks No indeed, they are not.
    
    \hamspeaks How comes it? doe they grow rusty?
    
    \rosinspeaks Nay, their indeauour keepes in the wonted pace; But there is Sir an ayrie of Children, little Yases, that crye out on the top of question; and are most tyrannically clap't for't: these are now the\\
    fashion, and so be-ratled the common Stages (so they call them) that many wearing Rapiers, are affraide of Goose-quils, and dare scarse come thither.
    
    \hamspeaks What are they Children? Who maintains 'em? How are they escoted? Will they pursue the Quality no longer then they can sing? Will they not say afterwards if they should grow themselues to common Players (as it is like most if their meanes are no better) their Writers do them wrong, to make them exclaim against their owne Succession.
    
    \rosinspeaks Faith thrre ha's bene much to do on both sides: and the Nation holds it no sinne, to tarre them to Controuersie. There was for a while, no mony bid for argument, vnlesse the Poet and the Player went to Cuffes in the Question.
    
    \hamspeaks Is't possible?
    
    \guilspeaks Oh there ha's beene much throwing about of Braines.
    
    \hamspeaks Do the Boyes carry it away?
    
    \rosinspeaks I that they do my Lord. Hercules \& his load too.
    
    \hamspeaks It is not strange: for mine Vnckle is King of Denmarke, and those that would make mowes at him while my Father liued; giue twenty, forty,, an hundred Ducates a peece, for his picture in Little. There is something in this more then Naturall, if Philosophie could finde it out.\\
    (\textit{Flourish for the players.})
    
    \guilspeaks There are the Players.
    
    \hamspeaks Gentlemen, you are welcom to Elsonower: your hands, come: The appurtenance of Welcome, is Fashion and Ceremony. Let me comply with you in the Garbe, lest my extent to the Players (which I tell you must shew fairely outward) should more appeare like entertainment then yours. You are welcome: but my Vnckle Father, and Aunt Mother are deceiu'd.
    
    \guilspeaks In what my deere Lord?
    
    \hamspeaks I am but mad North, North-West: when the Winde is Southerly, I know a Hawke from a Handsaw.
    
    \direct{Enter \pol.}
    
    \polspeaks Well be with you Gentlemen.
    
    \hamspeaks Hearke you \guil, and you too: at each eare a hearer: that great Baby you see there, is not yet out of his swathing clouts.
    
    \rosinspeaks Happily he's the second time come to them: for they say, an old man is twice a childe.
    
    \hamspeaks I will Prophesie. Hee comes to tell me of the Players. Mark it, you say right Sir: for a Monday morning 'twas so indeed.
    
    \polspeaks My Lord, I haue Newes to tell you.
    
    \hamspeaks My Lord, I haue Newes to tell you.\\
    When Rossius an Actor in Rome———
    
    \polspeaks The Actors are come hither my Lord.
    
    \hamspeaks Buzze, buzze.
    
    \polspeaks Vpon mine Honor.
    
    \hamspeaks Then can each Actor on his Asse———
    
    \polspeaks The best Actors in the world, either for Tragedie, Comedie, Historie, Pastorall: Pastoricall-Comicall-Historicall-Pastorall: Tragicall-Historicall: Tragicall-Comicali-Historicall-Pastorall: Scene indiuible, or Poem vnlimited. Seneca cannot be too heauy, nor Plautus too light, for the law of Writ, and the Liberty. These are the onely men.
    
    \hamspeaks O Iephta Iudge of Israel, what a Treasure had'st thou?
    
    \polspeaks What a Treasure had he, my Lord?
    
    \hamspeaks Why one faire Daughter, and no more,\\
    The which he loued passing well.
    
    \polspeaks Still on my Daughter.
    
    \hamspeaks Am I not i'th'right old Iephta?
    
    \polspeaks If you call me Iephta my Lord, I haue a daughter that I loue passing well.
    
    \hamspeaks Nay that followes not.
    
    \polspeaks What followes then, my Lord?
    
    \hamspeaks Why, As by lot, God wot: and then you know, It came to passe, as most like it was: The first rowe of the Pons Chanson will shew you more. For looke where my Abridgements come.\\
    (\textit{Enter foure or fiue Players.})\\
    Y'are welcome Masters, welcome all. I am glad to see thee well: Welcome good Friends. O my old Friend? °Thy face is valiant since I saw thee last: Com'st thou to beard me in Denmarke? What, my yong Lady and Mistris? Byrlady your Ladiship is neerer Heauen then when I saw you last, by the altitude of a Choppine. Pray God your voice like a peece of vncurrant Gold be not crack'd within the ring. Masters, you are all welcome:wee'l e'ne to't like French Faulconers, flie at any thing we see: wee'l haue a Speech straight. Come giue vs a tast of your quality: come, a passionate speech.
    
    
    \speaker{1. Player} What speech, my Lord?
    
    \hamspeaks I heard thee speak me a speech once, but it was neuer Acted: or if it was, not aboue once, for the Play I remember pleas'd not the Million, 'twas Cauiarie to the Generall: but it was (as I receiu'd it and others, whose iudgement in such matters, cried in the top of mine) an excellent Play; well digested in the Scœnes, set downe with as much modestie, as cunning. I remember one said, there was no Sallets in the lines, to make the matter sauoury; nor no matter in the phrase, that might indite the Author of affection, but cal'd it an honest method. One cheefe Speech in it, I cheefely lou'd, 'twas Æneas Tale to Dido, and thereabout of it especially, where he speaks of Priams slaughter. If it liue in your memory, begin at this Line, let me see, let me see: The rugged Pyrrhus like th'Hyrcanian Beast. It is not so: it begins with Pyrrhus\\
    The rugged Pyrrhus, he whose Sable Armes\\
    Blacke as his purpose, did the night resemble\\
    When he lay couched in the Ominous Horse,\\
    Hath now this dread and blacke Complexion smear'd\\
    With Heraldry more dismall: Head to foote\\
    Now is he to take Geulles, horridly Trick'd\\
    With blood of Fathers, Mothers, Daughters, Sonnes,\\
    Bak'd and impasted with the parching streets,\\
    That lend a tyrannous, and damned light\\
    To their vilde Murthers, roasted in wrath and fire,\\
    And thus o're-sized with coagulate gore,\\
    VVith eyes like Carbuncles, the hellish Pyrrhus\\
    Old Grandsire Priam seekes.
    
    \polspeaks Fore God, my Lord, well spoken, with good accent, and good discretion.
    
    \speaker{1. Player} Anon he findes him,\\
    Striking too short at Greekes. His anticke Sword,\\
    Rebellious to his Arme, lyes where it falles\\
    Repugnant to command: vnequall match,\\
    Pyrrhus at Priam driues, in Rage strikes wide:\\
    But with the whiffe and winde of his fell Sword,\\
    Th'vnnerued Father fals. Then senseless Illium,\\
    Seeming to feele his blow, with flaming top\\
    Stoopes to his Bace, and with a hideous crash\\
    Takes Prisoner Pyrrhus eare. For loe, his Sword\\
    Which was declining on the Milkie head\\
    Of Reuerend Priam, seem'd i'th Ayre to sticke:\\
    So as a painted Tyrant Pyrrhus stood,\\
    And like a Newtrall to his will and matter, did nothing.\\
    But as we often see against some storme,\\
    A silence in the Heauens, the Racke stand still,\\
    The bold windes speechlesse, and the Orbe below\\
    As hush as death: Anon the dreadfull Thunder\\
    Doth rend the Region. So after Pyrrhus pause,\\
    A ro wsed Vengeance sets him new a-worke,\\
    And neuer did the Cyclops hammers fall\\
    On Mars his Armours, forg'd for proofe Eterne,\\
    With lesse remorse then Pyrrhus bleeding sword\\
    Now falles on Priam.\\
    Out, out, thou Strumpet-Fortune, all you Gods,\\
    In generall Synod take away her power:\\
    Breake all the Spokes and Fallies from her wheele,\\
    And boule the round Naue downe the hill of Heauen,\\
    As low as to the Fiends.
    
    \polspeaks This is too long.
    
    \hamspeaks It shall to'th Barbars, with your beard. Prythee say on: He's for a Iigge, or a tale of Baudry, or hee sleepes. Say on; come to Hecuba.
    
    \speaker{1. Player} But who, O who, had seen the inobled Queen.
    
    \hamspeaks The inobled Queene?
    
    \polspeaks That's good: Inobled Queene is good.
    
    \speaker{1. Player} Run bare-foot vp and downe,\\
    Threatning the flame\\
    With Bisson Rheume: A clout about that head,\\
    Where late the Diadem stood, and for a Robe\\
    About her lanke and all ore-teamed Loines,\\
    A blanket in th'Alarum of feare caught vp.\\
    Who this had seene, with tongue in Venome steep'd,\\
    'Gainst Fortunes State, would Treason haue pronounc'd?\\
    But if the Gods themselues did see her then,\\
    When she saw Pyrrhus make malicious sport\\
    In mincing with his Sword her Husbands limbes,\\
    The instant Burst of Clamour that she made\\
    (Vnlesse things mortall moue them not at all)\\
    Would haue made milche the Burning eyes of Heauen,\\
    And passion in the Gods.
    
    \polspeaks Looke where he ha's not turn'd his colour, and ha's teares in's eyes. Pray you no more.
    
    \hamspeaks 'Tis well, Ile haue thee speake out the rest, soone. Good my Lord, will you see the Players wel bestow'd. Do ye heare, let them be well vs'd: for they are the Abstracts and breefe Chronicles of the time. After your death, you were better haue a bad Epitaph, then their ill report while you liued.
    
    \polspeaks My Lord, I will vse them according to their desart.
    
    \hamspeaks Gods bodykins man, better. Vse euerie man after his desart, and who should scape whipping: vse them after your own Honor and Dignity. The lesse they deserue, the more merit is in your bountie. Take them in.
    
    \polspeaks Come sirs. \hfill(\textit{Exit \pol.})
    
    \hamspeaks Follow him Friends: wee'l heare a play to morrow. Dost thou heare me old Friend, can you play the murther of Gonzago?
    
    Play.\\
    I my Lord.
    
    \hamspeaks Wee'l ha't to morrow night. You could for a need study a speech of some dosen or sixteene lines, which I would set downe, and insert in't? Could ye not?
    
    Play.\\
    I my Lord.
    
    \hamspeaks Very well. Follow that Lord, and looke you mock him not. My good Friends, Ile leaue you til night you are welcome to Elsonower?
    
    \rosinspeaks Good my Lord. \hfill({Exeunt.})
    
    \direct{Manet \ham.}
    
    \hamspeaks I so, God buy'ye: Now I am alone.\\
    Oh what a Rogue and Pesant slaue am I?\\
    Is it not monstrous that this Player heere,\\
    But in a Fixion, in a dreame of Passion,\\
    Could force his soule so to his whole conceit,\\
    That from her working, all his visage warm'd:\\
    Teares in his eyes, distraction in's Aspect,\\
    A broken voyce, and his whole Function suiting\\
    With Formes, to his Conceit? And all for nothing?\\
    For Hecuba?\\
    What's Hecuba to him, or he to Hecuba,\\
    That he should weepe for her? What would he doe,\\
    Had he the Motiue and the Cue for passion\\
    That I haue? He would drowne the Stage with teares,\\
    And cleaue the generall eare with horrid speech:\\
    Make mad the guilty, and apale the free,\\
    Confound the ignorant, and amaze indeed,\\
    The very faculty, of Eyes and Eares, Yet I,\\
    A dull and muddy-metled Rascall, peake\\
    Like Iohn a-dreames, vnpregnant of my cause,\\
    And can say nothing: No, not for a King,\\
    Vpon whose property, and most deere life,\\
    A damn'd defeate was made. Am I a Coward?\\
    Who calles me Villaine? breakes my pate a-croffe?\\
    Pluckes off my Beard, and blowes it in my face?\\
    Tweakes me by'th'Nose? giues me the Lye i'th'Throate,\\
    As deepe as to the Lungs? Who does me this?\\
    Ha? Why I should take it: for it cannot be,\\
    But I am Pigeon-Liuer'd, and lacke Gall\\
    To make Oppression bitter, or ere this,\\
    I should haue fatted all the Region Kites\\
    With this Slaues Offall, bloudy: a Bawdy villaine,\\
    Remorselesse, Treacherous, Letcherous, kindles villaine!\\
    Oh Vengeance!\\
    Who? What an Asse am I? I sure, this is most braue,\\
    That I, the Sonne of the Deere murthered,\\
    Prompted to my Reuenge by Heauen, and Hell,\\
    Must (like a Whore) vnpacke my heart with words,\\
    And fall a Cursing like a very Drab,\\
    A Scullion? Fye vpon't: Foh. About my Braine.\\
    I haue heard, that guilty Creatures sitting at a Play,\\
    Haue by the very cunning of the Scœne,\\
    Bene strooke so to the soule, that presently\\
    They haue proclaim'd their Malefactions.\\
    For Murther, though it haue no tongue, will speake\\
    With most myraculous Organ. Ile haue these Players,\\
    Play something like the murder of my Father,\\
    Before mine Vnkle. Ile obserue his lookes,\\
    Ile rent him to the quicke: If he but blench\\
    I know my course. The Spirit that I haue seene\\
    May be the Diuell, and the Diuel hath power\\
    T'assume a pleasing shape, yea and perhaps\\
    Out of my Weaknesse, and my Melancholly,\\
    As he is very potent with such Spirits,\\
    Abuses me to damne me. Ile haue grounds\\
    More Relatiue then this: The Play's the thing,\\
    Wherein Ile catch the Conscience of the King. \hfill(\textit{Exit.})
\end{drama*}

\act

\scene

\updatemarks

\StageDir{Enter \king, \queen, \pol, \ophe, \rosin, \guil, and Lords.}

\begin{drama*}
    \kingspeaks And can you by no drift of circumstance\\
    Get from him why he puts on this Confusion:\\
    Grating so harshly all his dayes of quiet\\
    With turbulent and dangerous Lunacy.
    
    \rosinspeaks He does confesse he feeles himselfe distracted,\\
    But from what cause he will by no meanes speake.
    
    \guilspeaks Nor do we finde him forward to be sounded,\\
    But with a crafty Madnesse keepes aloofe:\\
    When we would bring him on to some Confession\\
    Of his true state.
    
    \queenspeaks Did he receiue you well?
    
    \rosinspeaks Most like a Gentleman.
    
    \guilspeaks But with much forcing of his disposition.
    
    \rosinspeaks Niggard of question, but of our demands\\
    Most free in his reply.
    
    \queenspeaks Did you assay him to any pastime?
    
    \rosinspeaks Madam.it so fell out, that certain Players\\
    We ore-wrought on the way: of these we told him,\\
    And there did seeme in him a kinde of ioy\\
    To heare of it: They are about the Court,\\
    And (as I thinke) they haue already order\\
    This night to play before him.
    
    \polspeaks 'Tis most true:\\
    And he beseech'd me to intreate your Maiesties\\
    To heare, and see the matter.
    
    \kingspeaks With all my heart, and it doth much content me\\
    To heare him so inclin'd. Good Gentlemen,\\
    Giue him a further edge, and driue his purpose on\\
    To these delights.
    
    \rosinspeaks We shall my Lord.\hfill(\textit{Exeunt.})
    
    \kingspeaks Sweet \queen leaue vs too,\\
    For we haue closely sent for \ham hither,\\
    That he, as 'twere by accident, may there\\
    Affront Ophelia. Her Father, and my selfe (lawful espials)\\
    Will so bestow our selues, that seeing vnseene\\
    We may of their encounter frankely iudge,\\
    And gather by him, as he is behaued,\\
    If't be th'affliction of his loue, or no.\\
    That thus he suffers for.
    
    \queenspeaks I shall obey you,\\
    And for your part Ophelia, I do wish\\
    That your good Beauties be the happy cause\\
    Of Hamlets wildenesse: so shall I hope your Vertues\\
    Will bring him to his wonted way againe,\\
    To both your Honors.
    
    \ophespeaks Madam, I wish it may.
    
    \polspeaks \ophe, walke you heere. Gracious so please ye\\
    We will bestow our selues: Reade on this booke,\\
    That shew of such an exercise may colour\\
    Your lonelinesse. We are oft too blame in this,\\
    'Tis too much prou'd, that with Deuotions visage,\\
    And pious Action, we do surge o're\\
    The diuell himselfe.
    
    \kingspeaks Oh'tis true:\\
    How smart a lash that speech doth giue my Conscience?\\
    The Harlots Cheeke beautied with plaist'ring Art\\
    Is not more vgly to the thing that helpes it,\\
    Then is my deede, to my most painted word.\\
    Oh heauie burthen!
    
    \polspeaks I heare him comming, let's withdraw my Lord. \hfill(\textit{Exeunt.}
    
    \direct{Enter \ham.}
    
    \hamspeaks To be, or not to be, that is the Question:\\
    Whether 'tis Nobler in the minde to suffer\\
    The Slings and Arrowes of outragious Fortune,\\
    Or to take Armes against a Sea of troubles,\\
    And by opposing end them: to dye, to sleepe\\
    No more; and by a sleepe, to say we end\\
    The Heart-ake, and the thousand Naturall shockes\\
    That Flesh is heyre too? 'Tis a consummation\\
    Deuoutly to be wish'd. To dye to sleepe,\\
    To sleepe, perchance to Dreame; I, there's the rub,\\
    For in that sleepe of death, what dreames may come,\\
    When we haue shufflel'd off this mortall coile,\\
    Must giue vs pawse. There's the respect\\
    That makes Calamity of so long life:\\
    For who would beare the Whips and Scornes of time,\\
    The Oppressors wrong, the poore mans Contumely,\\
    The pangs of dispriz'd Loue, the Lawes delay,\\
    The infolence of Office, and the Spurnes\\
    That patient merit of the vnworthy takes,\\
    When he himselfe might his Quietus make\\
    With a bare Bodkin? Who would these Fardles beare\\
    To grunt and sweat vnder a weary life,\\
    But that the dread of something after death,\\
    The vndiscouered Countrey, from whose Borne\\
    No Traueller returnes, Puzels the will,\\
    And makes vs rather beare those illes we haue,\\
    Then flye to others that we know not of.\\
    Thus Conscience does make Cowards of vs all,\\
    And thus the Natiue hew of Resolution\\
    Is sicklied o're, with the pale cast of Thought,\\
    And enterprizes of great pith and moment,\\
    With this regard their Currants turne away,\\
    And loose the name of Action. Soft you now,\\
    The faire Ophelia? Nimph, in thy Orizons\\
    Be all my sinnes remembred.
    
    \ophespeaks Good my Lord,\\
    How does your Honor for this many a day?
    
    \hamspeaks I humbly thanke you: well, well, well.
    
    \ophespeaks My Lord, I haue Remembrances of yours,\\
    That I haue longed long to re-deliuer.\\
    I pray you now, receiue them.
    
    \hamspeaks No, no, I neuer gaue you ought.
    
    \ophespeaks My honor'd Lord, I know right well you did,\\
    And with them words of so sweet breath compos'd,\\
    As made the things more rich, then perfume left:\\
    Take these againe, for to the Noble minde\\
    Rich gifts wax poore, when giuers proue vnkinde.\\
    There my Lord.
    
    \hamspeaks Ha, ha: Are you honest?
    
    \ophespeaks My Lord.
    
    \hamspeaks Are you faire?
    
    \ophespeaks What meanes your Lordship?
    
    \hamspeaks That if you be honest and faire, your Honesty\\
    should admit no discourse to your Beautie.
    
    \ophespeaks Could Beautie my Lord, haue better Comerce then your Honestie?
    
    \hamspeaks I trulie: for the power of Beautie, will sooner transforme Honestie from what it is, to a Bawd, then the force of Honestie can translate Beautie into his likenesse. This was sometime a Paradox, but now the time giues it proofe. I did loue you once.
    
    \ophespeaks Indeed my Lord, you made me beleeue so.
    
    \hamspeaks You should not haue beleeued me. For vertue cannot so innocculate our old stocke, but we shall rellish of it. I loued you not.
    
    \ophespeaks I was the more deceiued.
    
    \hamspeaks Get thee to a Nunnerie. Why would'st thou be a breeder of Sinners? I am my selfe indifferent honest, but yet I could accuse me of such things, that it were better my Mother had not borne me. I am very prowd, reuengefull, Ambitious, with more offences at my becke, then I haue thoughts to put them in imagination, to giue them shape, or time to acte them in. What should such Fellows as I do crawling between heaven and Earth. We are arrant knaues all, beleeue none of vs. Goe thy wayes to a Nunnery. Where's your Father?
    
    \ophespeaks At home, my Lord.
    
    \hamspeaks Let the doores be shut vpon him, that he may play the Foole no way, but in's owne house. Farewell.
    
    \ophespeaks O helpe him, you sweet Heauens.
    
    \hamspeaks If thou doest Marry, Ile giue thee this Plague for, thy Dowrie. Be thou as chast as Ice, as pure as Snow, thou shalt not escape Calumny. Get thee to a Nunnery. Go, Farewell. Or if thou wilt needs Marry, marry a fool: for Wise mem know well enough, what monsters you make of them. To a Nunnery go, and quickly too. Farwell.
    
    \ophespeaks O heauenly Powers, restore him.
    
    \hamspeaks I haue heard of your pratlings too wel enough. God has giuen you one pace, and you make your selfe an other: you gidge, you amble, and you lispe, and nickname Gods creatures, and make your Wantonnesse, your Ignorance. Go too, Ile no more on't, it hath made me mad, I say, we will haue no more Marriages. Those that are married already, all but one shall liue, the rest shall keep as they are. To a Nunnery, go.\\
    (\textit{Exit Hamlet.})
    
    \ophespeaks O what a Noble minde is heere o're-throwne?\\
    The Courtiers, Soldiers, Schollers: Eye, tongue, sword,\\
    Th'expectansie and Rose of the faire State,\\
    The glasse of Fashion, and the mould of Forme,\\
    Th'obseru'd of all Obseruers, quite, quite downe.\\
    Haue I of Ladies most deiect and wretched,\\
    That suck'd the Honie of his Musicke Vowes:\\
    Now see that Noble, and most Soueraigne Reason,\\
    Like sweet Bels iangled out of tune, and harsh,\\
    That vnmatch'd Forme and Feature of blowne youth,\\
    Blasted with extasie. Oh, woe is me,\\
    T'haue seene what I haue seene: see what I see.
    
    \direct{Enter \king, and \pol.}
    
    \kingspeaks Loue? His affections do not that way tend,\\
    Nor what he spake, though it lack'd Forme a little,\\
    Was not like Madnesse. There's something in his soule?\\
    O're which his Melancholly sits on brood,\\
    And I do doubt the hatch, and the disclose\\
    Will be some danger, which to preuent\\
    I haue in quicke determination\\
    Thus set it downe. He shall with speed to England\\
    For the demand of our neglected Tribute:\\
    Haply the Seas and Countries different\\
    With variable Obiects, shall expell\\
    This something setled matter in his heart:\\
    Whereon his Braines still beating, puts him thus\\
    From fashion of himselfe. What thinke you on't?
    
    \polspeaks It shall do well. But yet do I beleeue\\
    The Origin and Commencement of this greefe\\
    Sprung from neglected loue. How now Ophelia?\\
    You neede not tell vs, what Lord Hamlet saide,\\
    We heard it all. My Lord, do as you please,\\
    But if you hold it fit after the Play,\\
    Let his Queene Mother all alone intreat him\\
    To shew his Greefes: let her be round with him,\\
    And Ile be plac'd so, please you in the eare\\
    Of all their conference. If she finde him not,\\
    To England send him: Or confine him where\\
    Your wisedome best shall thinke.
    
    \kingspeaks It shall be so:\\
    Madnesse in great Ones, must not vnwatch'd go. \hfill(\textit{Exeunt.})
\end{drama*}

\scene

\updatemarks

\StageDir{Enter \ham, and two or three of the Players.}

\begin{drama*}
    \hamspeaks Speake the Speech I pray you, as I pronounc'd it to you trippingly on the Tongue; But if you mouth it, as many of your Players do, I had as liue the Town-Cryer had spoke my Lines: Nor do not saw the Ayre too much your hand thus, but vse all gently; for in the verie Torrent, Tempest, and (as I may say) the Whirle-winde of Passion, you must acquire and beget a Temperance that may giue it Smoothnesse. O it offends mee to the Soule, to see a robustious Pery-wig-pated Fellow, teare a Passion to tatters, to verie ragges, to split the eares of the Groundlings: who (for the most part) are capeable of nothing, but inexplicable dumbe shewes, \& noise: I could haue such a Fellow whipt for o're-doing Termagant: it out-Herod's Herod. Pray you auoid it.
    
    \speaker{Player} I warrant your Honor.
    
    \hamspeaks Be not too tame neyther: but let your owne Discretion be your Tutor. Sute the Action to the Word, the Word to the Action, with this speciall obseruance: That you ore-stop not the modestie of Nature; for any thing so ouer-done, is (frō)from the purpose of Playing, whose end both at the first and now, was and is, to hold as 'twer the Mirrour vp to Nature; to shew Vertue her owne Feature, Scorne her owne Image, and the verie Age and Bodie of the Time, his forme and pressure. Now, this ouer-done, or come tardie off, though it make the vnskilfull laugh, cannot but make the Iudicious greeue; The censure of the which One, must in your allowance o're way a whole Theater of Others. Oh, there bee Players that I haue seene Play, and heard others praise, and that highly (not to speake it prophanely) that neyther hauing the accent of Christians, nor the gate of Christian, Pagan, or Norman, haue so strutted and bellowed, that I haue thought some of Natures Iouerney-men had made men, and not made them well, they imitated Humanity so abhominably.
    
    \speaker{Player} I hope we haue reform'd that indifferently with vs, Sir.
    
    \hamspeaks O reforme it altogether. And let those that play your Clownes, speake no more then is set downe for them. For there be of them, that will themselues laugh, to set on some quantitie of barren Spectators to laugh too, though in the meane time, some necessary Question of the Play be then to be considered: that's Villanous, \& shewes a most pittifull Ambition in the Foole that vses it. Go make you readie. \hfill(\textit{Exit Players.})\\
    \mbox{}\\
    (\textit{Enter \pol, \rosin, and \guil.})\\
    \mbox{}\\
    How now my Lord,\\
    Will the King heare this peece of Worke?
    
    \polspeaks And the Queene too, and that presently.
    
    \hamspeaks Bid the players make hast. \hfill(\textit{Exit \pol.})\\
    Will you two helpe to hasten them?
    
    \speaker{\rosin and \guil} We will my Lord. Exeunt.
    
    \direct{Enter \hor.}
    
    \hamspeaks What hoa, \hor?
    
    \horspeaks Heere sweet Lord, at your Seruice.
    
    \hamspeaks Horatio, thou art eene as just a man\\
    As ere my Conuersation coap'd withall.
    
    \horspeaks O my deere Lord.
    
    \hamspeaks Nay, do not thinke I flatter:\\
    For what aduancement may I hope from thee,\\
    That no Reuennew hast, but thy good spirits\\
    To feed \& cloath thee. Why shold the poor be flatter'd?\\
    No, let the Candied tongue, like absurd pompe,\\
    And crooke the pregnant Hindges of the knee,\\
    Where thrift may follow faining? Dost thou heare,\\
    Since my deere Soule was Mistris of my choyse,\\
    And could of men distinguish, her election\\
    Hath seal'd thee for her selfe. For thou hast bene\\
    As one in suffering all, that suffers nothing.\\
    A man that Fortunes buffets, and Rewards\\
    Hath 'tane with equall Thankes. And blest are those,\\
    Whose Blood and Iudgement are so well co-mingled,\\
    That they are not a Pipe for Fortunes finger,\\
    To sound what stop she please. Giue me that man,\\
    That is not Passions Slaue, and I will weare him\\
    In my hearts Core: I, in my Heart of heart,\\
    As I do thee. Something too much of this.\\
    There is a Play to night before the King.\\
    One Scœne of it comes neere the Circumstance\\
    Which I haue told thee, of my Fathers death.\\
    I prythee, when thou see'st that Acte a-foot,\\
    Euen with the verie Comment of my Soule\\
    Obserue mine Vnkle: If his occulted guilt,\\
    Do not it selfe vnkennell in one speech,\\
    It is a damned Ghost that we haue seene:\\
    And my Imaginations are as foule\\
    As Vulcans Stythe. Giue him needfull note,\\
    For I mine eyes will riuet to his Face:\\
    And after we will both our iudgements ioyne,\\
    To censure of his seeming.
    
    \horspeaks Well my Lord.\\
    If he steale ought the whil'st this Play is Playing,\\
    And scape detecting, I will pay the Theft.
    
    % typo in wikisource (Guidensterne)
    \direct{Enter \king, \queen, \pol, \ophe, \rosin, \guil, and other Lords attendant with his Guard carrying Torches. Danish March. Sound a Flourish.}
    
    \hamspeaks They are comming to the Play: I must be idle.\\
    Get you a place.
    
    \kingspeaks How fares our Cosin \ham?
    
    \hamspeaks Excellent I faith, of the Camelions dish: I eate the Ayre promise-cramm'd, you cannot feed Capons so.
    
    \kingspeaks I haue nothing with this answer \ham, these words are not mine.
    
    \hamspeaks No, nor mine. Now my Lord, you plaid once i'th'Vniuersity, you say?
    
    \polspeaks That I did my Lord, and was accounted a good Actor.
    
    \hamspeaks And what did you enact?
    
    \polspeaks I did enact Iulius Cæsar, I was kill'd i'th'Capitol:\\
    Brutus kill'd me.
    
    \hamspeaks It was a bruite part of him, to kill so Capitall a Calfe there. Be the Players ready?
    
    \rosinspeaks I my Lord, they stay vpon your patience.
    
    \queenspeaks Come hither my good \ham, sit by me.
    
    \hamspeaks No good Mother, here's Mettle more attractiue.
    
    \polspeaks Oh ho, do you marke that?
    
    \hamspeaks Ladie, shall I lye in your Lap?
    
    \ophespeaks No my Lord.
    
    \hamspeaks I meane, my Head vpon your Lap?
    
    \ophespeaks I my Lord.
    
    \hamspeaks Do you thinke I meant Country matters?
    
    \ophespeaks I thinke nothing, my Lord.
    
    \hamspeaks That's a faire thought to ly between Maids legs
    
    \ophespeaks What is my Lord?
    
    \hamspeaks Nothing.
    
    \ophespeaks You are merrie, my Lord?
    
    \hamspeaks Who I?
    
    \ophespeaks I my Lord.
    
    \hamspeaks Oh God, your onely Iigge-maker: what should a man do, but be merrie. For looke you how cheereful ly my Mother lookes, and my Father dyed within's two Houres.
    
    \ophespeaks Nay,'tis twice two moneths, my Lord.
    
    \hamspeaks So long? Nay then let the Diuel weare blacke, for Ile haue a suite of Sables. Oh Heauens! dye two moneths ago, and not forgotten yet? Then there's hope, a great mans Memorie, may out-liue his life halfe a yeare: But byrlady he must builde Churches then: or else shall he suffer not thinking on, with the Hoby-horsse, whose Epitaph is, For o, For o, the Hoby-horse is forgot.
    
    \direct{Hoboyes play. The dumbe shew enters.}
    \textit{Enter a King and Queene, very louingly; the Queene embracing him. She kneeles and makes shew of Protestation vnto him. He takes her vp, and [declines] his head vpon her neck:Layes him downe vpon a Banke of Flowers. She seeing him a-sleepe, leaues him. Anon comes in a Fellow, takes off his Crowne, kisses it, and powres poyson in the Kings eares, and Exits. The Queene returnes, findes the King dead, and makes passionate Action. The Poysoner, with some two or three Mutes comes in againe, seeming to lament with her. The dead body is carried away: The Poysoner Wooes the Queene with Gifts, she seemes loath and vnwilling awhile, but in the end, accepts his loue\\
    .}\\
    (\textit{Exeunt.})
    
    
    \ophespeaks What meanes this, my Lord?
    
    \hamspeaks Marry this is Miching Malicho, that meanes Mischeefe.
    
    \ophespeaks Belike this shew imports the Argument of the Play?
    
    \hamspeaks We shall know by these Fellowes: the Players cannot keepe counsell, they'l tell all.
    
    \ophespeaks Will they tell vs what this shew meant?
    
    \hamspeaks I, or any shew that you'l shew him. Bee not you asham'd to shew, hee'l not shame to tell you what it meanes.
    
    \ophespeaks You are naught, you are naught, Ile marke the Play.
    
    \direct{Enter Prologue.}
    \begin{center}\textit{
    For vs, and for our Tragedie,\\
    Heere stooping to your Clemencie:\\
    We begge your hearing Patientlie.\\
    }\end{center}
    
    \hamspeaks Is this a Prologue, or the Poesie of a Ring?
    
    \ophespeaks 'Tis briefe my Lord.
    
    \hamspeaks As Womans loue.
    
    \direct{Enter King and his Queene.}
    
    \playking Full thirtie times hath Phœbus Cart gon round,\\
    Neptunes salt Wash, and Tellus Orbed ground:\\
    And thirtie dozen Moones with borrowed sheene,\\
    About the World haue times twelue thirties beene,\\
    Since loue our hearts, and Hymen did our hands\\
    Vnite comutuall, in most sacred Bands.
    
    \playqueen So many iournies may the Sunne and Moone\\
    Make vs againe count o're, ere loue be done.\\
    But woe is me, you are so sicke of late,\\
    So farre from cheere, and from your forme state,\\
    That I distrust you: yet though I distrust,\\
    Discomfort you (my Lord) it nothing must:\\
    For womens Feare and Loue, holds quantitie,\\
    In neither ought, or in extremity:\\
    Now what my loue is, proofe hath made you know,\\
    And as my Loue is siz'd, my Feare is so.
    
    \playking Faith I must leaue thee Loue, and shortly too:\\
    My operant Powers my Functions leaue to do:\\
    And thou shalt liue in this faire world behinde,\\
    Honour'd, belou'd, and haply, one as kinde.\\
    For Husband shalt thou———
    
    \playqueen Oh confound the rest:\\
    Such Loue, must needs be Treason in my brest:\\
    In second Husband, let me be accurst,\\
    None wed the second. but who kill'd the first.
    
    \hamspeaks Wormwood, Wormwood.
    
    \playqueen The instances that second Marriage moue,\\
    Are base respects of Thrift, but none of Loue.\\
    A second time, I kill my Husband dead,\\
    When second Husband kisses me in Bed.
    
    \playking I do beleeue you. Think what now you speak:\\
    But what we do determine, oft we breake:\\
    Purpose is but the slaue to Memorie,\\
    Of violent Birth, but poore validitie:\\
    Which now like Fruite vnripe stickes on the Tree,\\
    But fall vnshaken, when they mellow bee.\\
    Most necessary 'tis, that we forget\\
    To pay our selues, what to our selues is debt:\\
    What to our selues in passion we propose,\\
    The passion ending, doth the purpose lose.\\
    The violence of other Greefe or Ioy,\\
    Their owne ennactors with themselues destroy:\\
    Where Ioy most Reuels, Greefe doth most lament;\\
    Greefe ioyes, Ioy greeues on slender accident.\\
    This world is not for aye, nor 'tis not strange\\
    That euen our Loues should with our Fortunes change.\\
    For 'tis a question left vs yet to proue,\\
    Whether Loue lead Fortune, or else Fortune Loue.\\
    The great man downe you marke his fauourites flies,\\
    The poore aduanc'd, makes Friends of Enemies:\\
    And hitherto doth Loue on Fortune tend,\\
    For who not needs, shall neuer lacke a Frend:\\
    And who in want a hollow Friend doth try,\\
    Directly seasons him his Enemie.\\
    But orderly to end, where I begun,\\
    Our Willes and Fates do so contrary run,\\
    That our Deuices still are ouerthrowne,\\
    Our thoughts are ours, their ends, none of our owne.\\
    So thinke thou wilt no second Husband wed.\\
    But die thy thoughts, when thy first Lord is dead.
    
    \playqueen Nor Earth to giue me food, not Heauen light,\\
    Sport and repose locke from me day and night:\\
    Each opposite that blankes the face of ioy,\\
    Meet what I would haue well, and it destroy:\\
    Both heere, and hence, pursue me lasting strife,\\
    If once a Widdow, euer I be Wife.
    
    \hamspeaks If she should breake it now.
    
    \playking 'Tis deepely sworne:\\
    Sweet, leaue me heere a while,\\
    My spirits grow dull, and faine I would beguile\\
    The tedious day with sleepe.
    
    \playqueen Sleepe rocke thy Braine, Sleepes\\
    And neuer come mischance betweene vs twaine. \hfill(\textit{Exit.})
    
    \hamspeaks Madam, how like you this Play?
    
    \queenspeaks The Lady protests to much me thinkes.
    
    \hamspeaks Oh but shee'l keepe her word.
    
    \kingspeaks Haue you heard the Argument, is there no Offence in't?
    
    \hamspeaks No, no, they do but iest, poyson in iest, no Offence i'th'world.
    
    \kingspeaks What do you call the Play?
    
    \hamspeaks The Mouse-trap: Marry how? Tropically: This Play is the Image of a murder done in Vienna: Gonzago is the Dukes name, his wife Baptista: you shall see anon: 'tis a knauish peece of worke: But what o'that? Your Maiestie, and wee that haue free soules, it touches vs not: let the gall'd iade winch: our withers are vnrung.\\
    (\textit{Enter Lucianus.})\\
    This is one Lucianus nephew to the King.
    
    \ophespeaks You are a good Chorus, my Lord.
    
    \hamspeaks I could interpret betweene you and your loue: if I could see the Puppets dallying.
    
    \ophespeaks You are keene my Lord, you are keene.
    
    \hamspeaks It would cost you a groaning, to take off my edge.
    
    \ophespeaks Still better and worse.
    
    \hamspeaks So you mistake Husbands.\\
    Begin Murderer. Pox, leaue thy damnable Faces, and begin. Come, the croaking Rauen doth bellow for Reuenge.
    
    \playlucianus Thoughts blacke, hands apt,\\
    Drugges fit, and Time agreeing:\\
    Confederate season, else, no Creature seeing:\\
    Thou mixture ranke, of Midnight Weeds collected,\\
    With Hecats Ban, thrice blasted, thrice infected,\\
    Thy naturall Magicke, and dire propertie,\\
    On wholsome life, vsurpe immediately.\\
    (\textit{Powres the poyson in his eares.})
    
    \hamspeaks He poysons him i'th'Garden for's estate: His name's Gonzago: the Story is extant and writ in choyce Italian. You shall see anon how the Murtherer gets the loue of Gonzago's wife.
    
    \ophespeaks The King rises.
    
    \hamspeaks What, frighted with false fire.
    
    \queenspeaks How fares my Lord?
    
    \polspeaks Giue o're the Play.
    
    \kingspeaks Giue me some Light. Away.
    
    \speaker{All} Lights, Lights, Lights. \hfill(\textit{Exeunt.})
    
    \direct{Manet \ham \& \hor.}
    
    \hamspeaks Why let the strucken Deere go weepe,\\
    The Hart vngalled play:\\
    For some must watch, while some must sleepe;\\
    So runnes the world away.\\
    Would not this Sir, and a Forrest of Feathers, if the rest of my Fortunes turne Turke with me; with two Prouniciall Roses on my rac'd Shooes, get me a Fellowship in a crie of Players sir.
    
    \horspeaks Halfe a share.
    
    \hamspeaks A whole one I,\\
    For thou dost know: Oh Damon deere,\\
    This Realme dismantled was of Ioue himselfe,\\
    And now reignes heere.\\
    A verie verie Paiocke.
    
    \horspeaks You might haue Rim'd.
    
    \hamspeaks Oh good \hor, Ile take the Ghosts word for a thousand pound. Did'st perceiue?
    
    \horspeaks Verie well my Lord.
    
    \hamspeaks Vpon the talke of the poysoning?
    
    \horspeaks I did verie well note him.
    
    \direct{Enter \rosin and \guil.}
    
    \hamspeaks Oh, ha? Come some Musick. Come ye Recorders:\\
    For if the King like not the Comedie,\\
    Why then belike he likes it not perdie.\\
    Come some Musicke.
    
    \guilspeaks Good my Lord, vouchsafe me a word with you.
    
    \hamspeaks Sir a whole History.
    
    \guilspeaks The King, sir.
    
    \hamspeaks I sir, what of him?
    
    \guilspeaks Is in his retyrement, maruellous distemper'd.
    
    \hamspeaks With drinke Sir?
    
    \guilspeaks No my Lord, rather with choller.
    
    \hamspeaks Your wisedome should shew it selfe more richer, to signifie this to his Doctor: for for me to put him to his Purgation, would perhaps plundge him into farre more Choller.
    
    \guilspeaks Good my Lord put your discourse into some frame, and start not so wildely from my affayre.
    
    \hamspeaks I am tame Sir, pronounce.
    
    \guilspeaks The Queene your Mother, in most great affliction of spirit, hath sent me to you.
    
    \hamspeaks You are welcome.
    
    \guilspeaks Nay, good my Lord, this courtesie is not of the right breed. If it shall please you to make me a wholsome answer, I will doe your Mothers command'ment: if not, your pardon, and my returne shall bee the end of my Businesse.
    
    \hamspeaks Sir, I cannot.
    
    \guilspeaks What, my Lord?
    
    \hamspeaks Make you a wholsome answere: my wits diseas'd. But sir, such answers as I can make, you shal command: or rather you say, my Mother: therfore no more but to the matter. My Mother you say.
    
    \rosinspeaks Then thus she sayes: your behauior hath stroke her into amazement, and admiration.
    
    \hamspeaks Oh wonderfull Sonne, that can so astonish a Mother. But is there no sequell at the heeles of this Mothers admiration?
    
    \rosinspeaks She desires to speake with you in her Closset, ere you go to bed.
    
    \hamspeaks We shall obey, were she ten times our Mother.\\
    Haue you any further Trade with vs?
    
    \rosinspeaks My Lord, you once did loue me.
    
    \hamspeaks So I do still, by these pickers and stealers.
    
    \rosinspeaks Good my Lord, what is your cause of distemper? You do freely barre the doore of your owne Libertie, if you deny your greefes to your Friend.
    
    \hamspeaks Sir I lacke Aduancement.
    
    \rosinspeaks How can that be, when you haue the voyce of the King himselfe, for your Succession in Denmarke?
    
    \hamspeaks I, but while the grasse growes, the Prouerbe is something musty.\\
    (\textit{Enter one with a Recorder.})\\
    O the Recorder. Let me see, to withdraw with you, why do you go about to recouer the winde of mee, as if you would driue me into a toyle?
    
    \guilspeaks O my Lord, if my Dutie be too bold, my loue is too vnmannerly.
    
    \hamspeaks I do not well vnderstand that. Will you play vpon this Pipe?
    
    \guilspeaks My Lord, I cannot.
    
    \hamspeaks I pray you.
    
    \guilspeaks Beleeue me, I cannot.
    
    \hamspeaks I do beseech you.
    
    \guilspeaks I know no touch of it, my Lord.
    
    \hamspeaks 'Tis as easie as lying: gouerne these Ventiges with your finger and thumbe, giue it breath with your mouth, and it will discourse most excellent Musicke. Looke you, these are the stoppes.
    
    \guilspeaks But these cannot I command to any vtterance of hermony, I haue not the skill.
    
    \hamspeaks Why looke you now, how vnworthy a thing you make of me: you would play vpon mee; you would seeme to know my stops: you would pluck out the heart of my Mysterie; you would sound mee from my lowest Note, to the top of my Compasse: and there is much Musicke, excellent Voice, in this little Organe, yet cannot you make it. Why do you thinke, that I am easier to bee plaid on, then a Pipe? Call me what Instrument you will, though you can fret me, you cannot play vpon me, God blesse you Sir.
    
    \direct{Enter \pol.}
    
    \polspeaks My Lord; the Queene would speak with you, and presently.
    
    \hamspeaks Do you see that Clowd? that's almost in shape like a Camell.
    
    \polspeaks By'th'Misse, and it's like a Camell indeed.
    
    \hamspeaks Me thinkes it is like a Weazell.
    
    \polspeaks It is back'd like a Weazell.
    
    \hamspeaks Or like a Whale?
    
    \polspeaks Verie like a Whale.
    
    \hamspeaks Then will I come to my Mother, by and by:\\
    They foole me to the top of my bent.\\
    I will come by and by.
    
    \polspeaks I will say so. \hfill(\textit{Exit.})
    
    \hamspeaks By and by, is easily said. Leaue me Friends:\\
    'Tis now the verie witching time of night,\\
    When Churchyards yawne, and Hell it selfe breaths out\\
    Contagion to this world. Now could I drink hot blood,\\
    And do such bitter businesse as the day\\
    Would quake to looke on. Soft now, to my Mother:\\
    Oh Heart, loose not thy Nature; let not euer\\
    The Soule of Nero, enter this firme bosome:\\
    Let me be cruell, not vnnaturall,\\
    I will speake Daggers to her, but vse none:\\
    My Tongue and Soule in this be Hypocrites.\\
    How in my words someuer she be shent,\\
    To giue them Seales, neuer my Soule consent.
\end{drama*}

\scene

\updatemarks

\StageDir{Enter \king, \rosin, and \guil.}

\begin{drama*}
    \kingspeaks I like him not, nor stands it safe with vs,\\
    To let his madnesse range. Therefore prepare you,\\
    I your Commission will forthwith dispatch,\\
    And he to England shall along with you:\\
    The termes of our estate, may not endure\\
    Hazard so dangerous as doth hourely grow\\
    Out of his Lunacies.
    
    \guilspeaks We will our selues prouide:\\
    Most holie and Religious feare it is\\
    To keepe those many many bodies safe\\
    That liue and feede vpon your Maiestie.
    
    \rosinspeaks The single\\
    And peculiar life is bound\\
    With all the strength and Armour of the minde,\\
    To keepe it selfe from noyance: but much more,\\
    That Spirit, vpon whose spirit depends and rests\\
    The liues of many, the cease of Maiestie\\
    Dies not alone; but like a Gulfe doth draw\\
    What's neere it, with it. It is a massie wheele\\
    Fixt on the Somnet of the highest Mount,\\
    To whose huge Spoakes, ten thousand lesser things\\
    Are mortiz'd and adioyn'd: which when it falles,\\
    Each small annexment, pettie consequence\\
    Attends the boystrous Ruine. Neuer alone\\
    Did the King sighe, but with a generall grone.
    
    \kingspeaks Arme you, I pray you to this speedie Voyage;\\
    For we will Fetters put vpon this feare,\\
    Which now goes too free-footed.
    
    \speaker{\rosin and \guil} We will haste vs. \hfill(\textit{Exeunt Gent.})
    
    \direct{Enter \pol.}
    
    \polspeaks My Lord, he's going to his Mothers Closset:\\
    Behinde the Arras Ile conuey my selfe\\
    To heare the Processe. Ile warrant shee'l tax him home,\\
    And as you said, and wisely was it said,\\
    'Tis meete that some more audience then a Mother,\\
    Since Nature makes them partiall, should o're-heare\\
    The speech of vantage. Fare you well my Liege,\\
    Ile call vpon you ere you go to bed,\\
    And tell you what I know.
    
    \kingspeaks Thankes deere my Lord.\\
    Oh my offence is ranke, it smels to heauen.\\
    It hath the primall eldest curse vpon't,\\
    A Brothers murther. Pray can I not,\\
    Though inclination be as sharpe as will:\\
    My stronger guilt, defeats my strong intent,\\
    And like a man to double businesse bound,\\
    I stand in pause where I shall first begin,\\
    And both neglect; what if this cursed hand\\
    Were thicker then it selfe with Brothers blood,\\
    Is there not Raine enough in the sweet Heauens\\
    To wash it white as Snow? Whereto serues mercy,\\
    But to confront the visage of Offence?\\
    And what's in Prayer, but this two-fold force,\\
    To be fore-stalled ere we come to fall,\\
    Or pardon'd being downe? Then Ile looke vp,\\
    My fault is past. But oh, what forme of Prayer\\
    Can serue my turne? Forgiue me my foule Murther:\\
    That cannot be, since I am still possest\\
    Of those effects for which I did the Murther.\\
    My Crowne, mine owne Ambition, and my Queene:\\
    May one be pardon'd, and retaine th'offence?\\
    In the corrupted currants of this world,\\
    Offences gilded hand may shoue by Iustice,\\
    And oft 'tis seene, the wicked prize it selfe\\
    Buyes out the Law; but 'tis not so aboue,\\
    There is no shuffling, there the Action lyes\\
    In his true Nature, and we our selues compell'd\\
    Euen to the teeth and forehead of our faults,\\
    To giue in euidence. What then? What rests?\\
    Try what Repentance can. What can it not?\\
    Yet what can it, when one cannot repent?\\
    Oh wretched state! Oh bosome, blacke as death!\\
    Oh limed soule, that strugling to be free,\\
    Art more ingag'd: Helpe Angels, make assay:\\
    Bow stubborne knees, and heart with strings of Steele,\\
    Be soft as sinewes of the new-borne Babe,\\
    All may be well.
    
    \direct{Enter \ham.}
    
    \hamspeaks Now might I do it pat, now he is praying,\\
    And now Ile doo't, and so he goes to Heauen,\\
    And so am I reueng'd: that would be scann'd,\\
    A Villaine killes my Father, and for that\\
    I his foule Sonne, do this same Villaine send\\
    To heauen. Oh this is hyre and Sallery, not Reuenge.\\
    He tooke my Father grossely, full of bread,\\
    With all his Crimes broad blowne, as fresh as May,\\
    And how his Audit stands, who knowes, saue Heauen:\\
    But in our circumstance and course of thought\\
    'Tis heauie with him; and am I then reueng'd,\\
    To take him in the purging of his Soule,\\
    When he is fit and season'd for his passage? No.\\
    Vp Sword, and know thou a more horrid hent\\
    When he is drunke asleepe: or in his Rage,\\
    Or in th'incestuous pleasure of his bed,\\
    At gaming, swearing, or about some acte\\
    That ha's no rellish of Saluation in't,\\
    Then trip him, that his heeles may kicke at Heauen,\\
    And that his Soule may be as damn'd and blacke\\
    As Hell, whereto it goes. My Mother stayes,\\
    This Physicke but prolongs thy sickly dayes. \hfill(\textit{Exit.})
    
    \kingspeaks My words flye vp, my thoughts remain below,\\
    Words without thoughts, neuer to Heauen go. \hfill(\textit{Exit.})
\end{drama*}

\scene

\updatemarks

\StageDir{Enter \queen and \pol.}

\begin{drama*}
    \polspeaks He will come straight:\\
    Looke you lay home to him,\\
    Tell him his prankes haue been too broad to beare with,\\
    And that your Grace hath [screen'd], and stoode betweene\\
    Much heate, and him. Ile silence me e'ene heere:\\
    Pray you be round with him.
    
    \speaker{\ham within.} Mother, mother, mother.
    
    \queenspeaks Ile warrant you, feare me not.\\
    Withdraw, I heare him comming.
    
    \direct{Enter \ham.}
    
    \hamspeaks Now Mother, what's the matter?
    
    \queenspeaks \ham, thou hast thy Father much offended.
    
    \hamspeaks Mother, you haue my Father much offended.
    
    \queenspeaks Come, come, you answer with an idle tongue.
    
    \hamspeaks Go, go, you question with an idle tongue.
    
    \queenspeaks Why how now \ham?
    
    \hamspeaks Whats the matter now?
    
    \queenspeaks Haue you forgot me?
    
    \hamspeaks No by the Rood, not so:\\
    You are the Queene, your Husbands Brothers wife,\\
    But would you were not so. You are my Mother.
    
    \queenspeaks Nay, then Ile set those to you that can speake.
    
    \hamspeaks Come, come, and sit you downe, you shall not boudge:\\
    You go not till I set you vp a glasse,\\
    Where you may see the inmost part of you?
    
    \queenspeaks What Wilt thou do? thou wilt not murther me?\\
    Helpe, helpe, hoa.
    
    \polspeaks What hoa, helpe, helpe, helpe.
    
    \hamspeaks How now, a Rat? dead for a Ducate, dead.
    
    \polspeaks Oh I am slaine. \hfill(\textit{Killes \pol.})
    
    \queenspeaks Oh me, what hast thou done?
    
    \hamspeaks Nay I know not, is it he King?
    
    \queenspeaks Oh what a rash, and bloody deed is this?
    
    \hamspeaks A bloody deed, almost as bad good Mother,\\
    As kill a King, and marrie with his Brother.
    
    \queenspeaks As kill a King?
    
    \hamspeaks I Lady, 'twas my word.\\
    Thou wretched, rash, intruding foole farewell,\\
    I tooke thee for thy Betters, take thy Fortune,\\
    Thou find'st to be too busie, is some danger.\\
    Leaue wringing of your hands, peace, sit you downe,\\
    And let me wring your heart, for so I shall\\
    If it be made of penetrable stuffe;\\
    If damned Custome haue not braz'd it so,\\
    That it is proofe and bulwarke against Sense.
    
    \queenspeaks What haue I done, that thou dar'st wag thy tong,\\
    In noise so rude against me?
    
    \hamspeaks Such an Act\\
    That blurres the grace and blush of Modestie,\\
    Cals Vertue Hypocrite, takes off the Rose\\
    From the faire forehead of an innocent loue,\\
    And makes a blister there. Makes marriage vowes\\
    As false as Dicers Oathes. Oh such a deed,\\
    As from the body of Contraction pluckes\\
    The very soule, and sweete Religion makes\\
    A rapsidie of words. Heauens face doth glow,\\
    Yea this solidity and compound masse,\\
    With tristfull visage as against the doome,\\
    Is thought-sicke at the act.
    
    \queenspeaks Aye me; what act; that roares so lowd, \& thunders in the Index.
    
    \hamspeaks Looke heere vpon this Picture, and on this,\\
    The counterfet presentment of two Brothers:\\
    See what a grace was seated on his Brow,\\
    Hyperions curles, the front of Ioue himselfe,\\
    An eye like Mars, to threaten or command\\
    A Station, like the Heraland Mercurie\\
    New lighted on a heauen-kissing hill:\\
    A Combination, and a forme indeed,\\
    Where euery God did seeme to set his Seale,\\
    To giue the world assurance of a man.\\
    This was your Husband. Looke you now what followes.\\
    Heere is your Husband, like a Mildew'd eare\\
    Blasting his wholsom breath. Haue you eyes?\\
    Could you on this faire Mountaine leaue to feed,\\
    And batten on this Moore? Ha? Haue you eyes?\\
    You cannot call it Loue: For at your age,\\
    The hey-day in the blood is tame, it's humble,\\
    And waites vpon the Iudgement: and what Iudgement\\
    Would step from this, to this? What diuell was't,\\
    That thus hath cousend yon at hoodman-blinde?\\
    O Shame! where is thy Blush? Rebellious Hell,\\
    If thou canst mutine in a Matrons bones,\\
    To flaming youth, let Vertue be as waxe,\\
    And melt in her owne fire. Prodaime no shame,\\
    When the compulsiue Ardure giues the charge,\\
    Since Frost it selfe, as actiuely doth burne,\\
    As Reason panders Will.
    
    \queenspeaks O Hamlet, speake no more.\\
    Thou turn'st mine eyes into my very soule,\\
    And there I see such blacke and grained spots,\\
    As will not leaue their Tinct.
    
    \hamspeaks Nay, but to liue\\
    In the ranke sweat of an enseamed bed,\\
    Stew'd in Corruption; honying and making loue\\
    Ouer the nasty Stye.
    
    \queenspeaks Oh speake to me no more,\\
    These words like Daggers enter in mine eares.\\
    No more sweet Hamlet.
    
    \hamspeaks A Murderer, and a Villaine:\\
    A Slaue, that is not twentieth [part] the tythe\\
    Of your precedent Lord. A vice of Kings,\\
    A Cutpurse of the Empire and the Rule.\\
    That from a shelfe, the precious Diadem stole,\\
    And put it in his Pocket.
    
    \queenspeaks No more.
    
    \direct{Enter \gho.}
    
    \hamspeaks A King of shreds and patches.\\
    Saue me; and houer o're me with your wings\\
    You heauenly Guards. What would you gracious figure?
    
    \queenspeaks Alas he's mad.
    
    \hamspeaks Do you not come your tardy Sonne to chide,\\
    That laps't in Time and Passion, lets go by\\
    Th'important acting of your dread command? Oh say.
    
    \ghospeaks Do not forget: this Visitation\\
    Is but to whet thy almost blunted purpose.\\
    But looke, Amazement on thy Mother sits;\\
    O step betweene her, and her fighting Soule,\\
    Conceit in weakest bodies, strongest workes.\\
    Speake to her \ham.
    
    \hamspeaks How is it with you Lady?
    
    \queenspeaks Alas, how is't with you?\\
    That you bend your eye on vacancie,\\
    And with their corporall ayre do hold discourse.\\
    Forth at your eyes, your spirits wildely peepe,\\
    And as the sleeping Soldiours in th'Alarme,\\
    Your bedded haire, like life in excrements,\\
    Start vp, and stand an end. Oh gentle Sonne,\\
    Vpon the heate and flame of thy distemper\\
    Sprinkle coole patience. Whereon do you looke?
    
    \hamspeaks On him, on him: look you how pale he glares.\\
    His forme and cause conioyn'd, preaching to stones,\\
    Would make them capeable. Do not looke vpon me,\\
    Least with this pitteous action you conuert\\
    My sterne effects: then what I haue to do,\\
    Will want true colour; teares perchance for blood.
    
    \queenspeaks To who do you speake this?
    
    \hamspeaks Do you see nothing there?
    
    \queenspeaks Nothing at all, yet all that is I see.
    
    \hamspeaks Nor did you nothing heare?
    
    \queenspeaks No, nothing but our selues.
    
    \hamspeaks Why look you there: looke how it steals away:\\
    My Father in his habite, as he liued,\\
    Look where he goes euen now out at the Portall. \hfill(\textit{Exit.})
    
    \queenspeaks This is the very coynage of your Braine,\\
    This bodilesse Creation extasie is very cunning in.
    
    \hamspeaks Exctasie?\\
    My Pulse as yours doth temperately keepe time,\\
    And makes as healthful Musicke. it is not madnesse\\
    That I haue vttered; bring me to the Test\\
    And I the matter will re-word: which madnesse\\
    Would gamboll from. Mother, for loue of Grace,\\
    Lay not a flattering Vnction to your soule,\\
    That not your trespasse, but my madnesse speakes;\\
    It will but skin and filme the Vlcerous place,\\
    Whil'st ranke Corruption mining all within,\\
    Infects vnseene. Confesse your selfe to Heauen,\\
    Repent what's past, auoyd what is to come,\\
    And do not spred the Compost or the Weedes,\\
    To make them ranke. Forgiue me this my Vertue,\\
    For in the fatnesse of this pursie times,\\
    Vertue it selfe, of Vice must pardon begge,\\
    Yea courb, and woe, for leaue to do him good.
    
    \queenspeaks Oh \ham,\\
    Thou hast cleft my heart in twaine.
    
    \hamspeaks O throw away the worser past of it,\\
    And liue the purer with the other halfe.\\
    Good night, but go not to mine Vnkles bed,\\
    Assume a Vertue, if you haue it not, refraine to night,\\
    And that shall lend a kinde of easinesse\\
    To the next abstinence. Once more goodnight,\\
    And when you are desirous o be blest,\\
    Ile blessing begge of you. For this fame Lord,\\
    I do repent: but heauen hath pleased it so,\\
    To punish me with this, and this with me,\\
    That I must be their Scourge and Minister.\\
    I will bestow him, and will answer well\\
    The death I gaue him: so againe, good night,\\
    I must be cruell, onely to be kinde;\\
    Thus bad begins, and worse remaines behinde.
    
    \queenspeaks What shall I do?
    
    \hamspeaks Not this by no meanes that I bid you do:\\
    Let the blunt King tempt you againe to bed,\\
    Pinch Wanton on your cheeke, call you his Mouse,\\
    And let him for a paire of reechie kisses,\\
    Or padling in your necke with his damn'd Fingers,\\
    Make you to rauell all this matter out,\\
    That I essentially am not in madnesse,\\
    But made in craft. 'Twere good you let him know,\\
    For who that's but a Queene, faire, sober, wise,\\
    Would from a Paddocke, from a Bat, a Gibbe,\\
    Such deere concernings hide, Who would do so,\\
    No in despight of Sense and Secrecie,\\
    Vnpegge the Basket on the houses top:\\
    Let the Birds flye, and like the famous Ape\\
    To try Conclusions in the Basket, creepe\\
    And breake your owne necke downe.
    
    \queenspeaks Be thou assur'd, if words be made of breath,\\
    And breath of life: I haue no life to breath\\
    What thou hast saide to me.
    
    \hamspeaks I must to England, you know that?
    
    \queenspeaks Alacke I had forgot: 'Tis so concluded on.
    
    \hamspeaks This man shall set me packing:\\
    Ile lugge the Guts into the Neighbor roome,\\
    Mother goodnight. Indeede this Counsellor\\
    Is now most still, most secret, and most graue,\\
    Who was in life, a foolish prating Knaue.\\
    Come sir, to draw toward an end with you.\\
    Good night Mother. \hfill(\textit{Exit \ham tugging in \pol.})
\end{drama*}

\act

\scene

\updatemarks

\StageDir{Enter \king.}

\begin{drama*}
    \kingspeaks There's matters in these sighes.\\
    These profound heaues\\
    You must translate; Tis fit we vnderstand them.\\
    Where is your Sonne?
    
    \queenspeaks Ah my good Lord, what haue I seene to night?
    
    \kingspeaks What \queen? How do's \ham?
    
    \queenspeaks Mad as the Seas, and winde, when both contend\\
    Which is the Mightier in his lawless fit\\
    Behinde the Arras, hearing some thing stirre,\\
    He whips his Rapier out, and cries a Rat, a Rat,\\
    And in his brainish apprehension killes\\
    The vnseene good old man.
    
    \kingspeaks Oh heauy deed:\\
    It had bin so with vs had we beene there:\\
    His Liberty is full of threats to all,\\
    To you your selfe, to vs, to euery one.\\
    Alas, how shall this bloody deede be answered?\\
    It will be laide to vs, whose prouidence\\
    Should haue kept short, restrain'd, and out of haunt,\\
    This mad yong man. But so much was our loue,\\
    We would not vnderstand what was most fit,\\
    But like the Owner of a foule disease,\\
    To keepe it from divulging, let's it feede\\
    Euen on the pith of life. Where is he gone?
    
    \queenspeaks To draw apart the body he hath kild,\\
    O're whom his very madnesse like some Oare\\
    Among a Minerall of Mettels base\\
    Shewes it selfe pure. He weepes for what is done.
    
    \kingspeaks Oh \queen, come away:\\
    The Sun no sooner shall the Mountaines touch,\\
    But we will ship him hence, and this vilde deed,\\
    We must with all our Maiesty and Skill\\
    Both countenance, and excuse.\\
    (\textit{Enter \rosin \& \guil})\\
    Ho \guil:\\
    Friends both go ioyne you with some further ayde:\\
    Hamlet in madnesse hath Polonius slaine,\\
    And from his Mother Clossets hath he drag'd him.\\
    Go seeke him out, speake faire, and bring the body\\
    Into the Chappell. I pray you hast in this.\\
    (\textit{Exit Gent.})\\
    Come Gertrude, wee'l call vp our wisest friends,\\
    To let them know both what we meane to do,\\
    And what's vntimely done. Oh come away,\\
    My soule is full of discord and dismay. \hfill(\textit{Exeunt.})
\end{drama*}

\scene

\updatemarks

\StageDir{Enter \ham.}

\begin{drama*}
    \hamspeaks Safely stowed.
    
    \speaker{Gentlemen within.} Hamlet, Lord Hamlet.
    
    \hamspeaks What noise? Who cals on Hamlet?\\
    Oh heere they come.\\
    (\textit{Enter \rosin and \guil.})
    
    \rosinspeaks What haue you done my Lord with the dead body?
    
    \hamspeaks Compounded it with dust, whereto 'tis Kinne.
    
    \rosinspeaks Tell vs where 'tis, that we may take it thence,\\
    And beare it to the Chappell.
    
    \hamspeaks Do not beleeue it.
    
    \rosinspeaks Beleeue what?
    
    \hamspeaks That I can keepe your counsell, and not mine owne. Besides, to be demanded of a Spundge, what replication should be made by the Sonne of a King.
    
    \rosinspeaks Take you me for a Spundge, my Lord?
    
    \hamspeaks I sir, that sokes vp the Kings Countenance, his Rewards, his Authorities (but such Officers do the King best seruice in the end. He keepes them like an Ape in the corner of his iaw, first mouth'd to be last swallowed, when he needes what you haue glean'd, it is but squeezing you, and Spundge you shall be dry againe.
    
    \rosinspeaks I vnderstand you not my Lord.
    
    \hamspeaks I am glad of it: a knauish speech sleepes in a foolish eare.
    
    \rosinspeaks My Lord, you must tell vs where the body is, and go with vs to the King.
    
    \hamspeaks The body is with the King, but the King is not with the body. The King, is a thing ———
    
    \guilspeaks A thing my Lord?
    
    \hamspeaks Of nothing: bring me to him, hide Fox, and all after. \hfill(\textit{Exeunt.})
\end{drama*}

\scene

\updatemarks

\StageDir{Enter \king.}

\begin{drama*}
    \kingspeaks I haue sent to seeke him, and to find the bodie:\\
    How dangerous is it that this man goes loose:\\
    Yet must not we put the strong Law on him:\\
    Hee's loued of the distracted multitude,\\
    Who like not in their iudgement, but their eyes:\\
    And where 'tis so, th'Offenders scourge is weigh'd\\
    But neerer the offence: to beare all smooth, and euen,\\
    This sodaine sending him away, must seeme\\
    Deliberate pause, diseases desperate growne,\\
    By desperate appliance are releeued,\\
    Or not at all.\\
    (\textit{Enter Rosincrane.})\\
    How now? What hath befalne?
    
    \rosinspeaks Where the dead body is bestow'd my Lord,\\
    We cannot get from him.
    
    \kingspeaks But where is he?
    
    \rosinspeaks Without my Lord, guarded to know your pleasure.
    
    \kingspeaks Bring him before vs.
    
    \rosinspeaks Hoa, \guil? Bring in my Lord.
    
    \direct{Enter \ham and \guil.}
    
    \kingspeaks Now \ham, where's \pol?
    
    \hamspeaks At Supper.
    
    \kingspeaks At Supper? Where?
    
    \hamspeaks Not where he eats, but where he is eaten, a certaine conuocation of wormes are e'ne at him. Your worm is your onely Emperor for diet. We fat all creatures else to fat vs, and we fat our selfe for Magots. Your fat King, and your leane Begger is but variable seruice to dishes, but to one Table that's the end.
    
    \kingspeaks What dost thou meane by this?
    
    \hamspeaks Nothing but to shew you how a King may go a Progresse through the guts of a Begger.
    
    \kingspeaks Where is \pol.
    
    \hamspeaks In heauen, send thither to see. If your Messenger finde him not there, seeke him i'th other place your selfe: but indeed, if you finde him not this moneth, you shall nose him as you go vp the staires into the Lobby.
    
    \kingspeaks Go seeke him there.
    
    \hamspeaks He will stay till ye come.
    
    \kingspeaks \ham, this deed of thine, for thine especial safety\\
    Which we do tender, as we deerely greeue\\
    For that which thou hast done, must send thee hence\\
    With fierie Quicknesse. Therefore prepare thy selfe,\\
    The Barke is readie, and the winde at helpe,\\
    Th'Associates tend, and euery thing at bent\\
    For England.
    
    \hamspeaks For England?
    
    \kingspeaks I \ham.
    
    \hamspeaks Good.
    
    \kingspeaks So is it, if thou knew'st our purposes.
    
    \hamspeaks I see a Cherube that see's him: but come, for England. Farewell deere Mother.
    
    \kingspeaks Thy louing Father Hamlet.
    
    \hamspeaks My Mother: Father and Mother is man and wife: man \& wife is one flesh, and so my mother. Come, for England. \hfill(\textit{Exit.})
    
    \kingspeaks Follow him at foote,\\
    Tempt him with speed aboord:\\
    Delay it not, Ile haue him hence to night.\\
    Away, for euery thing is Seal'd and done\\
    That else leanes on th'Affaire, pray you make hast.\\
    And England, if my loue thou holdst at ought,\\
    As my great power thereof may giue thee sense,\\
    Since yet thy Cicatrice lookes raw and red\\
    After the Danish Sword, and thy free awe\\
    Payes homage to vs; thou maist not coldly set\\
    Our Soueraigne Processe, which imports at full\\
    By Letters coniuring to that effect\\
    The present death of Hamlet. Do it England,\\
    For like the Hecticke in my blood he rages,\\
    And thou must cure me: Till I know 'tis done,\\
    How ere my happes, my ioyes were ne're begun. \hfill(\textit{Exit.})
\end{drama*}

\scene

\updatemarks

\StageDir{Enter \for with an Armie.}

\begin{drama*}
    \forspeaks Go Captaine, from me greet the Danish King,\\
    Tell him that by his license, Fortinbras\\
    Claimes the conueyance of a promis'd March\\
    Ouer his Kingdome. You know the Rendeuous:\\
    If that his Maiesty would ought with vs,\\
    We shall expresse our dutie in his eye,\\
    And let him know so.

    \speaker{Captaine} I will doo't, my Lord.

    \forspeaks Go safely on. \hfill(\textit{Exit.})
\end{drama*}

\scene

\updatemarks

\StageDir{Enter \queen and \hor.}

\begin{drama*}
    \queenspeaks I will not speake with her.
    
    \horspeaks She is importunate, indeed distract her moode will needs be pittied.
    
    \queenspeaks What would she haue?
    
    \horspeaks She speakes much of her Father; saies she heares\\
    There's trickes i'th'world, and hems, and beats her heart,\\
    Spumes enuiously at Strawes, speakes things in doubt,\\
    That carry but halfe sense: Her speech is nothing,\\
    Yet the vnshaped vse of it doth moue\\
    The hearers to Collection; they ayme at it,\\
    And botch the words vp fit to their owne thoughts,\\
    Which as her winkes, and nods, and gestures yeeld them,\\
    Indeed would make one thinke there would be thought,\\
    Though nothing sure, yet much vnhappily.
    
    \queenspeaks 'Twere good she were spoken with,\\
    For she may strew dangerous coniectures\\
    In ill breeding minds. Let her come in.\\
    To my sicke soule (as sinnes true Nature is)\\
    Each toy seemes Prologue, to some great amisse,\\
    So full of Artlesse iealousie is guilt,\\
    It spill's it selfe, in fearing to be spilt.
    
    \direct{Enter \ophe distracted.}
    
    \ophespeaks Where is the beauteous Maiesty of Denmark.
    
    \queenspeaks How now \ophe?
    
    \ophespeaks How shonld I your true loue know from another one?\\
    By his Cockle hat and staffe, and his Sandal shoone.
    
    \queenspeaks Alas sweet Lady: what imports this Song?
    
    \ophespeaks Say you? Nay pray you marke.\\
    He is dead and gone Lady, he is dead and gone,\\
    At his head a grasse-greene Turfe, at his heeles a stone.
    
    \direct{Enter \king.}
    
    \queenspeaks Nay but \ophe.
    
    \ophespeaks Pray you marke.\\
    \hspace{2em}\parbox{19em}{White his Shrow'd as the Mountaine Snow.}
    
    \queenspeaks Alas, looke heere my Lord.
    
    \ophespeaks \hspace{2em}\parbox{20em}{Larded with sweet flowers:\\
    Which bewept to the graue did not go,\\
    With true-loue showres.}
    
    \kingspeaks How do ye, pretty Lady?
    
    \ophespeaks Well, God dil'd you. They say the Owle was a Bakers daughter. Lord, wee know what we are, but know not what we may be. God be at your Table.
    
    \kingspeaks Conceit vpon her Father.
    
    \ophespeaks Pray you let's haue no words of this: but when they aske you what it meanes, say you this:\\
    \textit{To morrow is S. Valentines day, all in the morning betime,\\
    And I a Maid at your Window, to be your Valentine.\\
    Then vp he rose, \& don'd his clothes, \& dupt the chamber dore,\\
    Let in the Maid, that out a Maid, neuer departed more.}
    
    \kingspeaks Pretty \ophe.
    
    \ophespeaks Indeed la? without an oath Ile make an end ont.\\
    \hspace{2em}\parbox{20em}{By gis, and by (S.)Saint Charity,\\
    Alacke, and fie for shame:\\
    Yong men wil doo't, if they come too't,\\
    By Cocke they are too blame.\\
    Quoth she before you tumbled me,\\
    Yon promis'd me to Wed:\\
    So would I ha done by yonder Sunne,\\
    And thou hadst not come to my bed.}
    
    \kingspeaks How long hath she bin this?
    
    \ophespeaks I hope all will be well. We must bee patient, but I cannot choose but weepe, to thinke they should lay him i'th'cold ground: My brother shall knowe of it, and so I thanke you for your good counsell. Come, my Coach: Goodnight Ladies: Goodnight sweet Ladies: Goodnight, goodnight.\hfill(\textit{Exit.})
    
    \kingspeaks Follow her close,\\
    Giue her good watch I pray you:\\
    Oh this is the poyson of deepe greefe, it springs\\
    All from her Fathers death. Oh \queen, \queen,\\
    When sorrowes comes, they come not single spies,\\
    But in Battaliaes. First, her Father slaine,\\
    Next your Sonne gone, and he most violent Author\\
    Of his owne iust remoue: the people muddied,\\
    Thicke and vnwholsome in their thoughts, and whispers\\
    For good \pol death; and we haue done but greenly\\
    In hugger mugger to interre him. Poore \ophe\\
    Diuided from her selfe, and her faire Iudgement,\\
    Without the which we are Pictures, or meere Beasts.\\
    Last, and as much containing as all these,\\
    Her Brother is in secret come from France,\\
    Keepes on his wonder, keepes himselfe in clouds,\\
    And wants not Buzzers to infect his eare\\
    With pestilent Speeches of his Fathers death,\\
    Where in necessitie of matter Beggard,\\
    Will nothing sticke our persons to Arraigne\\
    In eare and eare. O my deere Gertrude, this,\\
    Like to a murdering Peece in many places,\\
    Giues me superfluous death. \hfill(\textit{A Noise within.})
    
    \direct{Enter a Messenger.}
    
    \queenspeaks Alacke, what noyse is this?
    
    \kingspeaks Where are my Switzers?\\
    Let them guard the doore. What is the matter?
    
    \speaker{Messenger} Saue your selfe, my Lord.\\
    The Ocean (ouer-peering of his List)\\
    Eates not the Flats with more impittious haste\\
    Then young \laer, in a Riotous head,\\
    Ore-beares your Officers, the rabble call him Lord,\\
    And as the world were now but to begin,\\
    Antiquity forgot, Custome not knowne,\\
    The Ratifiers and props of euery word,\\
    They cry choose we? \laer shall be King,\\
    Caps, hands, and tongues, applaud it to the clouds,\\
    \laer shall be King, \laer King.
    
    \queenspeaks How cheerefully on the false Traile they cry.\\
    Oh this is Counter you false Danish Dogges.
    
    \direct{Noise within.\hfill Enter \laer.}
    
    \kingspeaks The doores are broke.
    
    \laerspeaks Where is the King, sirs? Stand you all without.
    
    All.\\
    No, let's come in.
    
    \laerspeaks I pray you giue me leaue.
    
    Al.\\
    We will, we will.
    
    \laerspeaks I thanke you: Keepe the doore.\\
    Oh thou vilde King, giue me my Father.
    
    \queenspeaks Calmely good \laer.
    
    \laerspeaks That drop of blood, that calmes\\
    Proclaimes me Bastard:\\
    Cries Cuckold to my Father, brands the Harlot\\
    Euen here between the chaste vnsmirched brow\\
    Of my true Mother.
    
    \kingspeaks What is the cause \laer,\\
    That thy Rebellion lookes so Gyant-like?\\
    Let him go Gertrude: Do not feare our person:\\
    There's such Diuinity doth hedge a King,\\
    That Treason can but peepe to what it would,\\
    Acts little of his will. Tell me \laer,\\
    Why thou art thus Incenst? Let him go \queen.\\
    Speake man.
    
    \laerspeaks Where's my Father?
    
    \kingspeaks Dead.
    
    \queenspeaks But not by him.
    
    \kingspeaks Let him demand his fill.
    
    \laerspeaks How came he dead? Ile not be Iuggel'd with.\\
    To hell Allegeance: Vowes, to the blackest diuell.\\
    Conscience and Grace, to the profoundest Pit.\\
    I dare Damnation: to this point I stand,\\
    That both the worlds I giue to negligence,\\
    Let come what comes: onely Ile be reueng'd\\
    Most throughly for my Father.
    
    \kingspeaks Who shall stay you?
    
    \laerspeaks My Will, not all the world,\\
    And for, my meanes, Ile husband them so well,\\
    They shall go farre with little.
    
    \kingspeaks Good \laer:\\
    If you desire to know the certaintie\\
    Of your deere Fathers death, if writ in your reuenge,\\
    That Soop-stake you will draw both Friend and Foe,\\
    Winner and Looser.
    
    \laerspeaks None but his Enemies.
    
    \kingspeaks Will you know them then.
    
    \laerspeaks To his good Friends, thus wide Ile ope my Armes:\\
    And like the kinde Life-rend'ring Politician,\\
    Repast them with my blood.
    
    \kingspeaks Why now you speake\\
    Like a good Childe, and a true Gentleman.\\
    That I am guiltlesse of your Fathers death,\\
    And am most sensible in greefe for it,\\
    It shall as leuell to your Iudgement pierce\\
    As day do's to your eye.
    
    \direct{A noise within. Let her come in.}
    
    \direct{Enter \ophe.}
    
    \laerspeaks How now? what noise is that?\\
    Oh heate drie vp my Braines, teares seuen times salt,\\
    Burne out the Sence and Vertue of mine eye.\\
    By Heauen, thy madnesse shall be payed by waight,\\
    Till our Scale turnes the beame. Oh Rose of May,\\
    Deere Maid, kinde Sister, sweet \ophe:\\
    Oh Heauens, is't possible, a yong Maids wits,\\
    Should be as mortall as an old mans life?\\
    Nature is fine in Loue, and where 'tis fine,\\
    It sends some precious instance of it selfe\\
    After the thing it loues.
    
    \ophespeaks \hspace{2em}\parbox{20em}{They bore him Bare fac'd on the Beer,\\
    Hey non nony, nony, hey nony:\\
    And on his graue raines many a teare,\\
    Fare you well my Doue.}
    
    \laerspeaks Had'st thou thy wits, and did'st perswade Reuenge, it could not moue thus.
    
    \ophespeaks You must sing downe a-downe, and you call him a-downe-a. Oh, how the wheele becomes it? It is the false steward that stole his masters daughter.
    
    \laerspeaks This nothings more then matter.
    
    \ophespeaks There's Rosemary, that's for Remembraunce.\\
    Pray loue remember: and there is Paconcies, that's for\\
    Thoughts.
    
    \laerspeaks A document in madnesse, thoughts \& remembrance fitted.
    
    \ophespeaks There's Fennell for you, and Columbines: ther's Rew for you, and heere's some for me. Wee may call it Herbe Grace a Sundaies: Oh you must weare your Rew with a difference. There's a Daysie, I would giue you some Violets, but they wither'd all when my Father dyed: They say, he made a good end;\\
    \hspace{2em}\parbox{20em}{For bonny sweet Robin is all my ioy.}
    
    \laerspeaks Thought, and Affliction, Passion, Hell it selfe:\\
    She turnes to Fauour, and to prettinesse.
    
    \ophespeaks \hspace{4em}{And will he not come againe,\\
    And will he not come againe:\\
    No, no, he is dead, go to thy Death-bed,\\
    He neuer wil come againe.\\
    His Beard as white as Snow,\\
    All Flaxen was his Pole:\\
    He is gone, he is gone, and we cast away mone,\\
    Gramercy on his Soule.}\\
    And of all Christian Soules, I pray God.\\
    God buy ye. \hfill(\textit{Exeunt \ophe})
    
    \laerspeaks Do you see this, you Gods?
    
    \kingspeaks \laer, I must common with your greefe,\\
    Or you deny me right: go but apart,\\
    Make choice of whom your wisest Friends you will,\\
    And they shall heare and iudge 'twixt you and me;\\
    If by direct or by Colaterall hand\\
    They finde vs touch'd, we will our Kingdome giue,\\
    Our Crowne, our Life, and all that we call Ours\\
    To you in satisfaction. But if not,\\
    Be you content to lend your patience to vs,\\
    And we shall ioyntly labour with your soule\\
    To giue it due content.
    
    \laerspeaks Let this be so:\\
    His meanes of death, his obscure buriall;\\
    No Trophee, Sword, nor Hatchment o're his bones,\\
    No Noble rite, nor formall ostentation,\\
    Cry to be heard, as 'twere from Heauen to Earth,\\
    That I must call in question.
    
    \kingspeaks So you shall:\\
    And where th'offence is, let the great Axe fall.\\
    I pray you go with me. \hfill(\textit{Exeunt.})
\end{drama*}

\scene

\updatemarks

\StageDir{Enter \hor, with an Attendant.}

\begin{drama*}
    \horspeaks What are they that would speake with me?

    \speaker{Ser.} Saylors sir, they say they haue Letters for you.

    \horspeaks Let them come in,\\
    I do not know from what part of the world\\
    I should be greeted, if not from Lord Hamlet.\\

    \direct{Enter Saylor.}

    \speaker{Saylor} God blesse you Sir.

    \horspeaks  Let him blesse thee too.

    \speaker{Saylor} Hee shall Sir, and't please him. There's a Letter for you Sir: It comes from th'Ambassadours that was bound for England, if your name be Horatio, as I am let to know it is.

    \direct{Reads the Letter.}
    \lettrine{H}{oratio}, When thou shalt haue ouerlook'd this, giue these Fellowes some meanes to the King: They haue Letters for him. Ere we were two dayes old at Sea, a Pyrate of very Warlicke appointment gaue vs Chace. Finding our selues too slow of Saile, we put on a compelled Valour. In the Grapple, I boorded them: On the instant they got cleare of our Shippe, so I alone became their Prisoner. They haue dealt with mee, like Theeues of Mercy, but they knew what they did. I am to doe a good turne for them. Let the King haue the Letters I haue sent, and repaire thou to me with as much hast as thou wouldest flye death. I haue words to speake in your eare, will make thee dumbe, yet are they much too light for the bore of the Matter. These good Fellows will bring thee where I am. Rosincrance and Guildensterne, hold their course for England. Of them I haue much to tell thee, Farewell.\\
    \begin{flushright}
        \textit{He that thou knowest thine,\\
        \ham.}
    \end{flushright}
    Come, I will giue you way for these your Letters,\\
    And do't the speedier, that you may direct me\\
    To him from whom you brought them. \hfill(\textit{Exit.})
\end{drama*}

\scene

\updatemarks

\StageDir{Enter \king and \laer.}

\begin{drama*}
    \kingspeaks Now must your conscience my acquittance seal,\\
    And you must put me in your heart for Friend,\\
    Sith you haue heard, and with a knowing eare,\\
    That he which hath your Noble Father slaine,\\
    Pursued my life.
    
    \laerspeaks It well appeares. But tell me,\\
    Why you proceeded not against these feates,\\
    So crimefull, and so Capitall in Nature,\\
    As by your Safety, Wisedome, all things else,\\
    You mainly were stirr'd vp?
    
    \kingspeaks O for two speciall Reasons,\\
    Which may to you (perhaps) seeme much vnsinnowed,\\
    And yet to me they are strong. The Queen his Mother,\\
    Liues almost by his looks: and for my selfe,\\
    My Vertue or my Plague, be it either which,\\
    She's so coniunctiue to my life and soule;\\
    That as the starre moues not but in his Sphere,\\
    I could not but by her. The other Motiue,\\
    Why to a publike count I might not go,\\
    Is the great loue the generall gender beare him,\\
    Who dipping all his Faults in their affection,\\
    Would like the Spring that turneth Wood to Stone,\\
    Conuert his Gyues to Graces. So that my Arrowes\\
    Too slightly timbred for so loud a Winde,\\
    Would haue reuerted to my Bow againe,\\
    And not where I had arm'd them.
    
    \laerspeaks And so haue I a Noble Father lost,\\
    A Sister driuen into desperate tearmes,\\
    Who was (if praises may go backe againe)\\
    Stood Challenger on mount of all the Age\\
    For her perfections. But my reuenge will come.
    
    \kingspeaks Breake not your sleepes for that,\\
    You must not thinke\\
    That we are made of stuffe, so flat, and dull,\\
    That we can let our Beard be shooke with danger,\\
    And thinke it pastime. You shortly shall heare more,\\
    I lou'd your Father, and we loue our Selfe,\\
    And that I hope will teach you to imagine———\\
    \mbox{}\\
    (\textit{Enter a Messenger.})\\
    How now? What Newes?
    
    \speaker{Messenger} Letters my Lord from \ham. This to your Maiesty: this to the Queene.
    
    \kingspeaks From \ham? Who brought them?
    
    Mes.\\
    Saylors my Lord they say, I saw them not:\\
    They were giuen me by Claudio, he receiu'd them.
    
    \kingspeaks Laertes you shall heare them:\\
    Leaue vs. \hfill(\textit{Exit Messenger})\\
    \hspace{2em}\textit{High and Mighty, you shall know I am set naked on your Kingdome. To morrow shall I begge leaue to see your Kingly Eyes. When I shall (first asking your Pardon thereunto) recount th'Occasions of my sodaine, and more strange returne.}\\
    \hfill \ham.\\
    What should this meane? Are all the rest come backe?\\
    Or is it some abuse? Or no such thing?
    
    \laerspeaks Know you the hand?
    
    \kingspeaks 'Tis Hamlets Character, naked and in a Postscript here he sayes alone: Can you aduise me?
    
    \laerspeaks I'm lost in it my Lord; but let him come,\\
    It warmes the very sicknesse in my heart,\\
    That I shall liue and tell him to his teeth;\\
    Thus diddest thou.
    
    \kingspeaks If it be so Laertes, as how should it be so:\\
    How other wise will you be rul'd by me?
    
    \laerspeaks If so you'l not o'rerule me to a peace.
    
    \kingspeaks To thine owne peace: if he be now return'd,\\
    As checking at his Voyage, and that he meanes\\
    No more to vndertake it; I will worke him\\
    To an exploit now ripe in my Deuice,\\
    Vnder the which he shall not choose but fall;\\
    And for his death no winde of blame shall breath,\\
    But euen his Mother shall vncharge the practice,\\
    And call it accident: Some two Monthes hence\\
    Here was a Gentleman of Normandy.\\
    I'ue seene my selfe, and seru'd against the French,\\
    And they ran well on Horsebacke; but this Gallant
    
    Had witchcraft in't; he grew into his Seat,\\
    And to such wondrous doing brought his Horse,\\
    As had he beene encorps't and demy-Natur'd\\
    With the braue Beast, so farre he past my thought,\\
    That I in forgery of shapes and trickes,\\
    Come short of what he did.
    
    \laerspeaks A Norman was't?
    
    \kingspeaks A Norman.
    
    \laerspeaks Vpon my life Lamound.
    
    \kingspeaks The very same.
    
    \laerspeaks I know him well, he is the Brooch indeed,\\
    And Iemme of all our Nation.
    
    \kingspeaks Hee mad confession of you,\\
    And gaue you such a Masterly report,\\
    For Art and exercise in your defence;\\
    And for your Rapier most especially,\\
    That he cryed out, t'would be a sight indeed,\\
    If one could match you Sir. This report of his\\
    Did Hamlet so envenom with his Enuy,\\
    That he could nothing doe but wish and begge,\\
    Your sodaine comming ore to play with him;\\
    Now out of this.
    
    \laerspeaks Why out of this, my Lord?
    
    \kingspeaks Laertes was your Father deare to you?\\
    Or are you like the painting of a sorrow,\\
    A face without a heart?
    
    \laerspeaks Why aske you this?
    
    \kingspeaks Not that I thinke you did not loue your Father,\\
    But that I know Loue is begun by Time:\\
    And that I see in passages of proofe,\\
    Time qualifies the sparke and fire of it:\\
    Hamlet comes backe: what would you vndertake,\\
    To show your selfe your Fathers sonne indeed,\\
    More then in words?
    
    \laerspeaks To cut his throat i'th'Church.
    
    \kingspeaks No place indeed should murder Sancturize;\\
    Reuenge should haue no bounds: but good Laertes\\
    Will you doe this, keepe close within your Chamber,\\
    Hamlet return'd, shall know you are come home:\\
    Wee'l put on those shall praise your excellence,\\
    And set a double varnish on the fame\\
    The Frenchman gaue you, bring you in fine together,\\
    And wager on your heads, he being remisse,\\
    Most generous, and free from all contriuing,\\
    Will not peruse the Foiles? So that with ease,\\
    Or with a little shuffling, you may choose\\
    A Sword vnbaited, and in a passe of practice,\\
    Requit him for your Father.
    
    \laerspeaks I will doo't,\\
    And for that purpose Ile annoint my Sword:\\
    I bought an Vnction of a Mountebanke\\
    So mortall, I but dipt a knife in it,\\
    Where it drawes blood, no Cataplasme so rare,\\
    Collected from all Simples that haue Vertue\\
    Vnder the Moone, can saue the thing from death,\\
    That is but scratcht withall: Ile touch my point,\\
    With this contagion that if I gall him slightly,\\
    It may be death.
    
    \kingspeaks Let's further thinke of this,\\
    Weigh what conuenience both of time and meanes\\
    May fit vs to our shape, if this should faile;\\
    And that our drift looke through our bad performance,\\
    'Twere better not assaid; therefore this Project\\
    Should haue a backe or second, that might hold,\\
    If this should blast in proofe: Soft, let me see\\
    Wee'l make a solemne wager on your commings,\\
    I ha't: when in your motion you are hot and dry,\\
    As make your bowts more violent to the end,\\
    And that he cals for drinke; Ile haue prepar'd him\\
    A Challice for the nonce; whereon but sipping,\\
    If he by chance escape your venom'd stuck,\\
    Our purpose may hold there; how sweet Queene.
    
    \direct{Enter \queen.}
    
    \queenspeaks One woe doth tread vpon anothers heele,\\
    So fast they'l follow: your Sister's drown'd \laer.
    
    \laerspeaks Drown'd! O where?
    
    \queenspeaks There is a Willow growes aslant a Brooke,\\
    That shewes his hore leaues in the glassie streame:\\
    There with fantasticke Garlands did she come,\\
    Of Crow-flowers, Nettles, Daysies, and long Purples,\\
    That liberall Shepheards giue a grosser name;\\
    But our cold Maids doe Dead Mens Fingers call them:\\
    There on the pendant boughes, her Coronet weeds\\
    Clambring to hang; an enuious sliuer broke,\\
    When downe the weedy Trophies, and her selfe,\\
    Fell in the weeping Brooke, her cloathes spred wide,\\
    And Mermaid-like, a while they bore her vp,\\
    Which time she chaunted snatches of old tunes,\\
    As one incapable of her owne distresse,\\
    Or like a creature Natiue, and indued\\
    Vnto that Element: but long it could not be,\\
    Till that her garments, heauy with her drinke,\\
    Pul'd the poore wretch from her melodious buy,\\
    To muddy death.
    
    \laerspeaks Alas then, is she drown'd?
    
    \queenspeaks Drown'd, drown'd.
    
    \laerspeaks Too much of water hast thou poore Ophelia,\\
    And therefore I forbid my teares: but yet\\
    It is our tricke, Nature her custome holds,\\
    Let shame say what it will; when these are gone\\
    The woman will be out: Adue my Lord.\\
    I haue a speech of fire, that faine would blaze,\\
    But that this folly doubts it.\hfill(\textit{Exit.})
    
    \kingspeaks Let's follow, \queen:\\
    How much I had to doe to calme his rage?\\
    Now feare I this will giue it start againe;\\
    Therefore let's follow.\hfill(\textit{Exeunt.})
\end{drama*}

\act

\scene

\updatemarks

\StageDir{Enter two Clownes.}

\begin{drama*}
    \speaker{Clowen} Is she to bee buried in Christian buriall, that wilfully seeks her owne saluation?
    
    \speaker{Other} I tell thee she is, and therefore make her Graue straight. the Crowner hath sate on her, and finds it Christian buriall.
    
    \speaker{Clown} How can that be, vnlesse she drowned her selfe in her owne defence?
    
    \speaker{Other} Why 'tis found so.
    
    \speaker{Clown} It must be Se offendendo, it cannot bee else: for heere lies the point; If I drowne my selfe wittingly, it argues an Act: and an Act hath three branches. It is an Act to doe and to performe; argall she drown'd her selfe wittingly.
    
    \speaker{Other} Nay but heare you Goodman Deluer.
    
    \speaker{Clown} Giue me leaue; heere lies the water; good: heere stands the man; good: If the man goe to this water and drowne himselfe; it is will he nill he, he goes; marke you that? But if the water come to him \& drowne him; hee drownes not himselfe. Argall, hee that is not guilty of his owne death, shortens not his owne life.
    
    \speaker{Other} But is this law?
    
    \speaker{Clown} I marry is't, Crowners Quest Law.\\
    Will you ha the truth on't: if this had not beene a Gentlewoman, shee should haue beene buried out of Christian Buriall.
    
    \speaker{Clown} Why there thou say'st. And the more pitty that great folke should haue countenance in this world to drowne or hang themselves, more then their euen Christian. Come, my Spade; there is no ancient Gentlemen, but Gardiners, Ditchers and Graue-makers; they hold vp Adams Profession.
    
    \speaker{Other} Was he a Gentleman?
    
    \speaker{Clown} He was the first that euer bore Armes.
    
    \speaker{Other} Why he had none.
    
    \speaker{Clown} What, ar't a Heathen? how dost thou vnderstand the Scripture? the Scripture sayes Adam dig'd; could hee digge without Armes? Ile put another question to thee; if thou answerest me not to the purpose, confesse thy selfe———
    
    \speaker{Other} Go too.
    
    \speaker{Clown} What is he that builds stronger then either the Mason, the Shipwright, or the Carpenter?
    
    \speaker{Other} The Gallowes maker; for that Frame outliues a thousand Tenants.
    
    \speaker{Clown} I like thy wit well in good faith, the Gallowes does well; but how does it well? it does well to those that doe ill: now, thou dost ill to say the Gallowes is built stronger then the Church: Argall, the Gallowes may doe well to thee. Too't againe, Come.
    
    \speaker{Other} Who builds stronger then a Mason, a Shipwright, or a Carpenter?
    
    \speaker{Clown} I, tell me that, and vnyoake.
    
    \speaker{Other} Marry, now I can tell.
    
    \speaker{Clown} Too't.
    
    \speaker{Other} Masse, I cannot tell.
    
    \direct{Enter \ham and \hor a farre off.}
    
    \speaker{Clown} Cudgell thy braines no more about it; for your dull Asse will not mend his pace with beating; and when you are ask't this question next, say a Graue-maker: the Houses that he makes, lasts till Doomesday: go, get thee to Taughan, fetch me a stoupe of Liquor.\\
    \direct{Sings.\\
    In youth when I did loue, did loue,\\
    \hspace{1em}me thought it was very sweete:\\
    To contract O the time for a my behoue,\\
    \hspace{1em}O me thought there was nothing meete.}
    
    \hamspeaks Ha's this fellow no feeling of his businesse, that he sings at Graue-making?
    
    \horspeaks Custome hath made it in him a property of easinesse.
    
    \hamspeaks Tis ee'n so; the hand of little Imployment hath the daintier sense.
    
    \direct{Clowne sings.\\
    But Age with his stealing steps\\
    \hspace{1em}hath caught me in his clutch:\\
    And hath shipped me intill the Land,\\
    \hspace{1em}as if I had neuer beene such.}
    
    \hamspeaks That Scull had a tongue in it, and could sing once: how the knaue iowles it to th' grownd, as if it were Caines Iaw-bone, that did the first murther: It might be the Pate of a Polititian which this Asse o're Offices: one that could circumuent God, might it not?
    
    \horspeaks It might, my Lord.
    
    \hamspeaks Or of a Courtier, which could say, Good Morrow sweet Lord: how dost thou, good Lord? this might be my Lord such a one, that prais'd my Lord such a ones Horse, when he meant to begge it; might it not?
    
    \horspeaks I, my Lord.
    
    \hamspeaks Why ee'n so: and now my Lady Wormes, Chaplesse, and knockt about the Mazard with a Sextons Spade; heere's fine Reuolution, if wee had the tricke to fee't. Did these bones cost no more the breeding, but to play at Loggets with 'em? mine ake to thinke on't.
    
    \direct{Clowne sings.\\
    A Pickhaxe and a Spade, a Spade.\\
    \hspace{1em}for and a shrowding‐Sheete:\\
    O a Pit of Clay for to be made,\\
    \hspace{1em}for such a Guest is meete.}
    
    \hamspeaks There's another: why might not that bee the Scull of of a Lawyer? where be his Quiddits now? his Quillets? his Cases? his Tenures, and his Tricks? why doe's he suffer this rude knaue now to knocke him about the Sconce with a dirty Shouell, and will not tell him of his Action of Battery? hum. This fellow might be in's time a great buyer of Land, with his statutes, his Recognizances, his Fines, his double Vouchers, his Recoueries: Is this the fine of his Fines, and the recouery of his Recoueries, to haue his fine Pate full of fine Dirt? will his Vouchers vouch him no more of his Purchases, and double ones too, then the length and breadth of a paire of Indentures? the very Conueyances of his Lands will hardly lye in this Boxe; and must the Inheritor himselfe haue no more? ha?
    
    \horspeaks Not a iot more, my Lord.
    
    \hamspeaks Is not Parchment made of Sheep-skinnes?
    
    \horspeaks I my Lord, and of Calue-skinnes too.
    
    \hamspeaks They are Sheepe and Calues that seek out assurance in that. I will speake to this fellow; whose Graue's this Sir?
    
    \speaker{Clown} Mine Sir:\\
    \hspace{2em}\parbox{18em}{O a Pit of Clay for to be made,\\
    for such a Guest is meete.}
    
    \hamspeaks I thinke it be thine indeed: for thou liest in't.
    
    \speaker{Clown} You lye out on't Sir, and therefore it is not yours: for my part, I doe not lye in't; and yet it is mine.
    
    \hamspeaks Thou dost lye in't, to be in't and say 'tis thine: 'tis for the dead, not for the quicke, therefore thou lyest.
    
    \speaker{Clown} 'Tis a quicke lye Sir, 'twill away againe from me to you.
    
    \hamspeaks What man dost thou digge it for?
    
    \speaker{Clown} For no man Sir.
    
    \hamspeaks What woman then?
    
    \speaker{Clown} For none neither.
    
    \hamspeaks Who is to be buried in't?
    
    \speaker{Clown} One that was a woman Sir; but rest her Soule, shee's dead.
    
    \hamspeaks How absolute the knaue is? wee must speake by the Carde, or equiuocation will vndoe vs: by the Lord \hor, these three yeares I haue taken note of it, the Age is growne so picked, that the toe of the Pesant comes so neere the heeles of our Courtier, hee galls his Kibe. How long hast thou been a Graue-maker?
    
    \speaker{Clown} Of all the dayes i'th yeare, I came too't that day that our last King Hamlet o'recame \for.
    
    \hamspeaks How long is that since?
    
    \speaker{Clown} Cannot you tell that? euery foole can tell that: It was the very day, that young \ham was borne, hee that was mad, and sent into England.
    
    \hamspeaks I marry, why was he sent into England?
    
    \speaker{Clown} Why, because he was mad; hee shall recouer his wits there; or if he do not, it's no great matter there.
    
    \hamspeaks Why?
    
    \speaker{Clown} 'Twill not beseene in him, there the men are as mad as he.
    
    \hamspeaks How came he mad?
    
    \speaker{Clown} Very strangely they say.
    
    \hamspeaks How strangely?
    
    \speaker{Clown} Faith e'ene with loosing his wits.
    
    \hamspeaks Vpon what ground?
    
    \speaker{Clown} Why heere in Denmarke: I haue bin sixeteene heere, man and Boy thirty yeares.
    
    \hamspeaks How long will a man lie'ith'earth ere he rot?
    
    \speaker{Clown} Ifaith, if he be not rotten before he die (as we haue many pocky Coarses now adaies, that will scarce hold the laying in) he will last you some eight yeare, or nine yeare. A Tanner will last you nine year e.
    
    \hamspeaks Why he, more then another?
    
    \speaker{Clown} Why sir, his hide is so tan'd with his Trade, that he will keepe out water a great while. And your water, is a sore Decayer of your horson dead body. Heres a Scull now: this Scul, has laine in the earth three \& twenty years.
    
    \hamspeaks Whose was it?
    
    \speaker{Clown} A whorson mad Fellowes it was;\\
    Whose doe you think it was?
    
    \hamspeaks Nay, I know not.
    
    \speaker{Clown} A pestlence on him for a mad Rogue, a pou'rd a Flaggon of Renish on my head once. This same Scull Sir, this same Scull sir, was Yoricks Scull, the Kings Iester.
    
    \hamspeaks This?
    
    \speaker{Clown} E'ene that.
    
    \hamspeaks Let me see. Alas poore Yorick, I knew him \hor, a fellow of infinite Iest; of most excellent fancy, he hath borne me on his backe a thousand times. And how abhorred my Imagination is, my gorge rises at it. Heere hung those lipps, that I haue kist I know not how oft. VVhere be your Iibes now? Your Gambals? Your Songs? Your flashes of Merriment that were wont to set the Table on a Rore? No one now to mock your own Ieering? Quite chopfalne? Now get you to my Ladies Chamber, and tell her, let her paint an inch thicke, to this fauour she must come. Make her laugh at that: pry thee \hor tell me one thing.
    
    \horspeaks What's that my Lord?
    
    \hamspeaks Dost thou thinke Alexander lookt o'this fashion i'th'earth?
    
    \horspeaks E'ene so.
    
    \hamspeaks And smelt so? Puh.
    
    \horspeaks E'ene so, my Lord.
    
    \hamspeaks To what base vses we may returne \hor. Why may not Imagination trace the Noble dust of Alexander, till he find it stopping a bunghole.
    
    \horspeaks 'Twere to consider: to curiously to consider so.
    
    \hamspeaks No faith, not a iot. But to follow him thether with modestie enough, \& likeliehood to lead it; as thus. Alexander died: Alexander was buried: Alexander returneth into dust; the dust is earth; of earth we make Lome, and why of that Lome (whereto he was conuerted) might they not stopp a Beere-barrell?\\
    Imperiall Cæsar, dead and turn'd to clay,\\
    Might stop a hole to keepe the winde away.\\
    Oh, that that earth, which kept the world in awe,\\
    Should patch a Wall, t'expell the winters flaw.\\
    But soft, but soft, aside; heere comes the King.\\
    \mbox{}\\
    (\textit{Enter \king, \queen, \laer, and a Coffin, with Lords attendant.})\\
    \mbox{}\\
    The Queene, the Courtiers. Who is that they follow,\\
    And with such maimed rites? This doth betoken,\\
    The Coarse they follow, did with disperate hand,\\
    Fore do it owne life; 'twas fome Estate.\\
    Couch we a while, and mark.
    
    \laerspeaks What Cerimony else?
    
    \hamspeaks That is \laer, a very Noble youth: Marke.
    
    \laerspeaks What Cerimony else?
    
    \speaker{Priest} Her Obsequies haue bin as farre inlarg'd.\\
    As we haue warrantis, her death was doubtfull,\\
    And but that great Command, o're-swaies the order,\\
    She should in ground vnsanctified haue lodg'd,\\
    Till the last Trumpet. For charitable praier,\\
    Shardes, Flints, and Peebles, I should be throwne on her:\\
    Yet heere she is allowed her Virgin Rites,\\
    Her Maiden strewments, and the bringing home\\
    Of Bell and Buriall.
    
    \laerspeaks Must there no more be done?
    
    \speaker{Priest} No more be done:\\
    We should prophane the seruice of the dead,\\
    To sing sage Requiem, and such rest to her\\
    As to peace-parted Soules.
    
    \laerspeaks Lay her i'th'earth,\\
    And from her faire and vnpolluted flesh,\\
    May Violets spring. I tell thee (churlish Priest)\\
    A Ministring Angell shall my Sister be,\\
    When thou liest howling?
    
    \hamspeaks What, the faire \ophe?
    
    \queenspeaks Sweets, to the sweet farewell.\\
    I hop'd thou should'st haue bin my Hamlets wife:\\
    I thought thy Bride-bed to haue deckt (sweet Maid)\\
    And not t'haue strew'd thy Graue.
    
    \laerspeaks Oh terrible woer,\\
    Fall ten times trebble, on that cursed head\\
    Whose wicked deed, thy most Ingenious sence\\
    Depriu'd thee of. Hold off the earth a while,\\
    Till I haue caught her once more in mine armes:\hfill(\textit{Leaps in the graue.})\\
    Now pile your dust, vpon the quick, and dead,\\
    Till of this flat a Mountaine you haue made,\\
    To o're top old Pelion, or the skyish head\\
    Of blew Olympus.
    
    \hamspeaks What is he, whose griefes\\
    Beares such an Emphasis? whose phrase of Sorrow\\
    Coniure the wandrinig Starres, and makes them stand\\
    Like wonder-wounded hearers? This is I,\\
    \ham the Dane.
    
    \laerspeaks The deuill take thy soule.
    
    \hamspeaks Thou prai'st not well,\\
    I prythee take thy fingers from my throat;\\
    Sir though I am not Spleenatiue, and rash,\\
    Yet haue I fomething in me dangerous,\\
    Which let thy wisenesse feare. Away thy hand.
    
    \kingspeaks Pluck them asunder.
    
    \queenspeaks \ham, \ham.
    
    \speaker{Gen.} Good my Lord be quiet.
    
    \hamspeaks Why I will fight with him vppon this Theme,\\
    Vntill my eielids will no longer wag.
    
    \queenspeaks Oh my Sonne, what Theame?
    
    \hamspeaks I lou'd \ophe; fortie thousand Brothers\\
    Could not (with all there quanitie of Loue)\\
    Make vp my summe. What wilt thou do for her?
    
    \kingspeaks Oh he is mad \laer,
    
    \queenspeaks For loue of God forbeare him.
    
    \hamspeaks Come show me what thou'lt doe.\\
    Woo't weepe? Woo't fight? Woo't teare thy selfe?\\
    Woo't drinke vp Esile, eate a Crocodile?
    
    Ile doo't. Dost thou come heere to whine;\\
    To outface me with leaping in her Graue?\\
    Be buried quicke with her, and so will I.\\
    And if thou prate of Mountaines; let them throw\\
    Millions of Akers on vs; till our ground\\
    Sindging his pate against the burning Zone,\\
    Make Ossa like a wart. Nay, and thoul't mouth,\\
    Ile rant as well as thou.
    
    \kingspeaks This is meere Madnesse;\\
    And thus awhile the fit will worke on him:\\
    Anon as patient as the female Doue,\\
    When that her golden Cuplet are disclos'd;\\
    His silence will sit drooping.
    
    \hamspeaks Heare you Sir:\\
    What is the reason that you vse me thus?\\
    I loud' you euer; but it is no matter:\\
    Let Hercules himselfe doe what he may,\\
    The Cat will Mew, and Dogge will haue his day.\hfill(\textit{Exit.})
    
    \kingspeaks I pray you good \hor wait vpon him,\\
    Strengthen you patience in our last nights speech,\\
    Wee'l put the matter to the present push:\\
    Good \queen set some watch ouer your Sonne,\\
    This Graue shall haue a liuing Monument:\\
    An houre of quiet shortly shall we see;\\
    Till then, in patience our proceeding be.\hfill(\textit{Exeunt.})
\end{drama*}

\scene

\updatemarks

\StageDir{Enter \ham and \hor.}

\begin{drama*}
    \hamspeaks So much for this Sir; now let me see the other,\\
    You doe remember all the Circumstance.
    
    \horspeaks Remember it my Lord?
    
    \hamspeaks Sir, in my heart there was a kinde of fighting,\\
    That would not let me sleepe; me thought I lay\\
    Worse then the mutines in the Bilboes, rashly,\\
    (And praise be rashnesse for it) let vs know,\\
    Our indiscretion sometimes serues vs well,\\
    When our deare plots do paule, and that should teach vs,\\
    There's a Diuinity that shapes our ends,\\
    Rough-hew them how we will.
    
    \horspeaks That is most certaine.
    
    \hamspeaks Vp from my Cabin\\
    My sea-gowne scarft about me in the darke,\\
    Grop'd I to finde out them; had my desire,\\
    finger'd their Packet, and in fine, withdrew\\
    To mine owne roome againe, making so bold,\\
    (My feares forgetting manners) to vnseale\\
    Their grand Commission, where I found \hor,\\
    Oh royall knauery: An exact command,\\
    Larded with many seuerall sorts of reason;\\
    Importing Denmarks health, and Englands too,\\
    With hoo, such Bugges and Goblins in my life,\\
    that on the superuize no leasure bated,\\
    No not to stay the grinding of the Axe,\\
    My head should be struck off.
    
    \horspeaks Ist possible?
    
    \hamspeaks Here's the Commission, read it at more leysure:\\
    But wilt thou heare me how I did proceed?
    
    \horspeaks I beseech you.
    
    \hamspeaks Being thus benetted round with Villaines,\\
    Ere I could make a Prologue to my braines,\\
    They had begun the Play. I sate me downe,\\
    Deuis'd a new Commission, wrote it faire,\\
    I once did hold it as our Statists doe,\\
    A basenesse to write faire; and laboured much\\
    How to forget that learning: but Sir now,\\
    It did me Yeomans seruice: wilt thou know\\
    The effects of what I wrote?
    
    \horspeaks I, good my Lord.
    
    \hamspeaks An earnest Coniuration from the King,\\
    As England was his faithfull Tributary,\\
    As loue betweene them, as the Palme should flourish,\\
    As Peace should still her wheaten Garland weare,\\
    And stand a Comma 'tweene their amities,\\
    And many such like Assis of great charge,\\
    That on the view and know of these Contents,\\
    Without debatement further, more or lesse,\\
    He should the bearers put to sodaine death,\\
    Not shriuing time allowed.
    
    \horspeaks How was this seal'd?
    
    \hamspeaks Why, euen in that was Heauen ordinate;\\
    I had my fathers Signet in my Purse,\\
    Which was the Model of that Danish Seale:\\
    Folded the Writ vp in forme of the other,\\
    Subscrib'd it, gau't th'impression, plac't it safely,\\
    The changeling neuer knowne: Now, the next day\\
    Was our Sea fight, and what to this was sement,\\
    Thou know'st already.
    
    \horspeaks So \guil and \rosin, go too't.
    
    \hamspeaks Why man, they did make loue to this imployment\\
    They are not neere my Conscience; their debate\\
    Doth by their owne insinuation grow:\\
    'Tis dangerous, when the baser nature comes\\
    Betweene the passe, and fell incensed points\\
    Of mighty opposites.
    
    \horspeaks Why, what a King is this?
    
    \hamspeaks Does it not, thinkst thee, stand me now vpon\\
    He that hath kil'd my King, and whor'd my Mother,\\
    Pop t in betweene th'election and my hopes,\\
    Throwne out his Angle for my proper life,\\
    And with such coozenage; is't not perfect conscience,\\
    To quit him with this arme? And is't not to be damn'd\\
    To let this Canker of our nature come\\
    In further euill.
    
    \horspeaks It must be shortly knowne to him from England\\
    What is the issue of the businesse there.
    
    \hamspeaks It will be short,\\
    The interim's mine, and a mans life's no more\\
    Then to say one: but I am very sorry good \hor,\\
    That to \laer I forgot my selfe;\\
    For by the image of my Cause, I see\\
    The Portraiture of his; Ile count his fauours:\\
    But sure the brauery of his griefe did put me\\
    Into a Towring passion.
    
    \horspeaks Peace, who comes heere?
    
    \direct{Enter young \osr.}
    
    \osrspeaks Your Lordship is right welcome back to Denmarke.
    
    \hamspeaks I humbly thank you Sir, dost know this waterflie?
    
    \horspeaks No my good Lord.
    
    \hamspeaks Thy state is the more gracious; for'tis a vice to know him: he hath much Land, and fertile; let a Beast be Lord of Beasts, and his Crib shall stand at the Kings Messe; 'tis a Chowgh; but as I saw spacious in the possession of dirt.
    
    \osrspeaks Sweet Lord, if your friendship were at leysure, I should impart a thing to you from his Maiesty.
    
    \hamspeaks I will receiue it with all diligence of spirit; put your Bonet to his right vse, 'tis for the head.
    
    \osrspeaks I thanke your Lordship,'tis very hot.
    
    \hamspeaks No, beleeue mee 'tis very cold, the winde is Northerly.
    
    \osrspeaks It is indifferent cold my Lord indeed.
    
    \hamspeaks Mee thinkes it is very soultry, and hot for my Complexion.
    
    \osrspeaks Exceedingly, my Lord, it is very soutry, as 'twere I cannot tell how: but my Lord, his Maiesty bad me signifie to you, that he ha's laid a great wager on your head:\\
    Sir, this is the matter.
    
    \hamspeaks I beseech you remember.
    
    \osrspeaks Nay, in good faith, for mine ease in good faith: Sir, you are not ignorant of what excellence \laer is at his weapon.
    
    \hamspeaks What's his weapon?
    
    \osrspeaks Rapier and dagger.
    
    \hamspeaks That's two of his weapons; but well.
    
    \osrspeaks The sir King ha's wag'd with him six Barbary Horses, against the which he impon'd as I take it, sixe French Rapiers and Poniards, with their assignes, as Girdle, Hangers or so; three of the Carriages infaith are very deare to fancy, very responsiue to the hilts, most delicate carriages, and of very liberall conceit.
    
    \hamspeaks What call you the Carriages?
    
    \osrspeaks The Carriages Sir, are the hangers.
    
    \hamspeaks The phrase would bee more Germaine to the matter: If we could carry Cannon by our sides; I would it might be Hangers till then; but on sixe Barbary Horses against sixe French Swords: their Assignes, and three liberall conceited Carriages, that's the French but against the Danish; why is this impon'd as you call it?
    
    \osrspeaks The King Sir, bath laid that in a dozen passes betweene you and him, hee shall not exceed you three hits; He hath one twelue for mine, and that would come to imediate tryall, if your Lordship would vouchsafe the Answere.
    
    \hamspeaks How if I answere no?
    
    \osrspeaks I meane my Lord, the opposition of your person in tryall.
    
    \hamspeaks Sir, I will walke heere in the Hall; if it please his Maiestie, 'tis the breathing time of day with me; let the Foyles bee brought, the Gentleman willing, and the King hold his purpose; I will win for him if I can: if not, Ile gaine nothing but my shame, and the odde hits.
    
    \osrspeaks Shall I redeliuer you ee'n so?
    
    \hamspeaks To this effect Sir, after what flourish your nature will.
    
    \osrspeaks I commend my duty to your Lordship.
    
    \hamspeaks Yours, yours; hee does well to commend it himselfe, there are no tongues else for's tongue.
    
    \horspeaks This Lapwing runs away with the shell on his head.
    
    \hamspeaks He did Complie with his Dugge before hee suck't it: thus had he and mine more of the same Beauy that I know the drossie age dotes on; only got the tune of the time, and outward habite of encounter, a kinde of yesty collection, which carries them through \& through the most fond and winnowed opinions;and doe but blow them to their tryalls: the Bubbles are out.
    
    \horspeaks You will lose this wager, my Lord.
    
    \hamspeaks I doe not thinke so, since he went into France, I haue beene in continuall practice; I shall winne at the oddes: but thou wouldest not thinke how all heere about my heart: but it is no matter.
    
    \horspeaks Nay, good my Lord.
    
    \hamspeaks It is but foolery; but it is such a kinde of gain-giuing as would perhaps trouble a woman.
    
    \horspeaks If your minde dislike any thing, obey. I will forestall their repaire hither, and say you are not fit.
    
    \hamspeaks Not a whit, we defie Augury; there's a speciall\\
    Prouidence in the fall of a sparrow. If it be now,'tis not\\
    to come: if it bee not come, it will bee now: if it\\
    be not now; yet it will come; the readinesse is all, since no\\
    man ha's ought of what he leaues. What is't to leaue be\\
    times?
    
    \direct{Enter \king, \queen, \laer and Lords, with other Attendants with Foyles, and Gauntlets, a Table and Flagons of Wine on it.}
    
    \kingspeaks Come \ham, come, and take this hand from me.
    
    \hamspeaks Giue me your pardon Sir, I'ue done you wrong,\\
    But pardon't as you are a Gentleman.\\
    This presence knowes,\\
    And you must needs haue heard how I am punisht\\
    With sore distraction? What I haue done\\
    That might your nature honour, and exception\\
    Roughly awake, I heere proclaime was madnesse:\\
    Was't \ham wrong'd \laer? Neuer \ham.\\
    If \ham from himselfe be tane away:\\
    And when he's not himselfe, do's wrong \laer,\\
    Then \ham does it not, \ham denies it:\\
    Who does it then? His Madnesse? If't be so,\\
    \ham is of the Faction that is wrong'd,\\
    His madnesse is poore Hamlets Enemy.\\
    Sir, in this Audience,\\
    Let my disclaiming from a purpos'd euill,\\
    Free me so farre in your most generous thoughts,\\
    That I haue shot mine Arrow o're the house,\\
    And hurt my Mother.
    
    \laerspeaks I am satisfied in Nature,\\
    Whose motiue in this case should stirre me most\\
    To my Reuenge. But in my termes of Honor\\
    I stand aloofe, and will no reconcilement,\\
    Till by some elder Masters of knowne Honor,\\
    I haue a voyce, and president of peace\\
    To keepe my name vngorg'd. But till that time,\\
    I do receiue your offer'd loue like loue,\\
    And wil not wrong it.
    
    \hamspeaks I do embrace it freely,
    
    And will this Brothers wager frankely play.\\
    Giue vs the Foyles: Come on.
    
    \laerspeaks Come one for me.
    
    \hamspeaks Ile be your foile \laer, in mine ignorance,\\
    Your Skill shall like a starre i'th'darkest night,\\
    Sticke fiery off indeede.
    
    \laerspeaks You mocke me Sir.
    
    \hamspeaks No by this hand.
    
    \kingspeaks Giue them the Foyles yong \osr,\\
    Cousen \ham, you know the wager.
    
    \hamspeaks Verie well my Lord,\\
    Your Grace hath laide the oddes a'th'weaker side.
    
    \kingspeaks I do not feare it,\\
    I haue seene you both:\\
    But since he is better'd, we haue therefore oddes.
    
    \laerspeaks This is too heauy,\\
    Let me see another.
    
    \hamspeaks This likes me well,\\
    These Foyles haue all a length. \hfill(\textit{Prepare to play.})
    
    \osrspeaks I my good Lord.
    
    \kingspeaks Set me the Stopes of wine vpon that Table:\\
    If \ham giue the first, or second hit,\\
    Or quit in answer of the third exchange,\\
    Let all the Battlements their Ordinance fire,\\
    The King shal drinke to Hamlets better breath,\\
    And in the Cup an vnion shal he throw\\
    Richer then that, which foure successiue Kings\\
    In Denmarkes Crowne haue worne.\\
    Giue me the Cups,\\
    And let the Kettle to the Trumpets speake,\\
    The Trumpet to the Cannoneer without,\\
    The Cannons to the Heauens, the Heauen to Earth,\\
    Now the King drinkes to \ham. Come, begin,\\
    And you the Iudges beare a wary eye.
    
    \hamspeaks Come on sir.
    
    \laerspeaks Come on sir.
    
    \direct{They play.}
    
    \hamspeaks One.
    
    \laerspeaks No.
    
    \hamspeaks Iudgement.
    
    \osrspeaks A hit, a very palpable hit.
    
    \laerspeaks Well: againe.
    
    \kingspeaks Stay, giue me drinke.\\
    \ham, this Pearle is thine,\\
    Here's to thy health. Giue him the cup.\\
    (\textit{Trumpets sound, and shot goes off.})
    
    \hamspeaks Ile play this bout first, set by a-while.\\
    Come: Another hit; what say you?
    
    \laerspeaks A touch, a touch, I do confesse.
    
    \kingspeaks Our Sonne shall win.
    
    \queenspeaks He's fat, and scant of breath.\\
    Heere's a Napkin, rub thy browes,\\
    The Queene Carowses to thy fortune, \ham.
    
    \hamspeaks Good Madam.
    
    \kingspeaks \queen, do not drinke.
    
    \queenspeaks I will my Lord;\\
    I pray you pardon me.
    
    \kingspeaks It is the poyson'd Cup, it is too late.
    
    \hamspeaks I dare not drinke yet Madam,\\
    By and by.
    
    \queenspeaks Come, let me wipe thy face.
    
    \laerspeaks My Lord, Ile hit him now.
    
    \kingspeaks I do not thinke't.
    
    \laerspeaks And yet 'tis almost 'gainst my conscience.
    
    \hamspeaks Come, for the third.\\
    \laer, you but dally,\\
    I pray you passe with your best violence,\\
    I am affear'd you make a wanton of me.
    
    \laerspeaks Say you so? Come on. \hfill(\textit{Play.})
    
    \osrspeaks Nothing neither way.
    
    \laerspeaks Haue at you now.
    
    \direct{In scuffling they change Rapiers.}
    
    \kingspeaks Part them, they are incens'd.
    
    \hamspeaks Nay come, againe.
    
    \osrspeaks Looke to the Queene there hoa.
    
    \horspeaks They bleed on both sides. How is't my Lord?
    
    \osrspeaks How is't \laer?
    
    \laerspeaks Why as a Woodcocke\\
    To mine Sprindge, \osr,\\
    I am iustly kill'd with mine owne Treacherie.
    
    \hamspeaks How does the Queene?
    
    \kingspeaks She sounds to see them bleede.
    
    \queenspeaks No, no, the drinke, the drinke.\\
    Oh my deere \ham, the drinke, the drinke,\\
    I am poyson'd.
    
    \hamspeaks Oh Villany! How? Let the doore be lock'd.\\
    Treacherie, seeke it out.
    
    \laerspeaks it is heere \ham.\\
    \ham, thou art slaine,\\
    No Medicine in the world can do thee good.\\
    In thee, there is not halfe an houre of life;\\
    The Treacherous Instrument is in thy hand,\\
    Vnbated and envenom'd: the foule practise\\
    Hath turn'd it selfe on me. Loe, heere I lye,\\
    Neuer to rise againe: Thy Mothers poyson'd:\\
    I can no more, the King, the King's too blame.
    
    \hamspeaks The point envenom'd too,\\
    Then venome to thy worke.\hfill(\textit{Hurts the King.})
    
    \speaker{All} Treason, Treason.
    
    \kingspeaks O yet defend me Friends, I am but hurt.
    
    \hamspeaks Heere thou incestuous, murdrous,\\
    Damned Dane,\\
    Drinke off this Potion: Is thy Vnion heere?\\
    Follow my Mother. \hfill(\textit{King Dyes.})
    
    \laerspeaks He is iustly seru'd.\\
    It is a poyson temp'red by himselfe:\\
    Exchange forgiuenesse with me, Noble \ham;\\
    Mine and ray Fathers death come not vpon thee,\\
    Nor thine on me.\hfill(\textit{Dyes.})
    
    \hamspeaks Heauen make thee free of it, I follow thee.\\
    I am dead \hor, wretched Queene adiew,\\
    You that looke pale, and tremble at this chance,\\
    That are but Mutes or audience to this acte:\\
    Had I but time (as this fell Sergeant death\\
    Is strick'd in his Arrest) oh I could tell you.\\
    But let it be: \hor, I am dead,\\
    Thou liu'st, report me and my causes right\\
    To the vnsatisfied.
    
    \horspeaks Neuer beleeue it.\\
    I am more an Antike Roman then a Dane:\\
    Heere's yet some Liquor left.
    
    \hamspeaks As th'art a man, giue me the Cup.\\
    Let go, by Heauen Ile haue't.\\
    Oh good \hor, what a wounded name,\\
    (Things standing thus vnknowne) shall liue behind me.\\
    If thou did'st euer hold me in thy heart,\\
    Absent thee from felicitie a while,\\
    And in this harsh world draw thy breath in paine,\\
    To tell my Storie.\\
    (\textit{March afarre off, and shout within.})\\
    What warlike noyse is this?
    
    Enter \osr.
    
    \osrspeaks Yong \for, with conquest come from Poland\\
    To th' Ambassadors of England giues this warlike volly.
    
    \hamspeaks O I dye \hor:\\
    The potent poyson quite ore-crowes my spirit,\\
    I cannot liue to heare the Newes from England,'\\
    But I do prophesie th'election lights\\
    On \for, he ha's my dying voyce,\\
    So tell him with the occurrents more and lesse,\\
    Which haue solicited. The rest is silence, O, o, o, o.\hfil(\textit{Dyes.})
    
    \horspeaks Now cracke a Noble heart:\\
    Goodnight sweet Prince,\\
    And flights of Angels sing thee to thy rest,\\
    Why do's the Drumme come hither?
    
    \direct{Enter \for and English Ambassador, with Drumme, Colours, and Attendants.}
    
    
    \forspeaks Where is this sight?
    
    \horspeaks What is it ye would see;\\
    If ought of woe, or wonder, cease your search.
    
    \forspeaks His quarry cries on hauocke. Oh proud death,\\
    What feast is toward in thine eternall Cell.\\
    That thou so many Princes, at a shoote,\\
    So bloodily hast strooke.
    
    \speaker{English Ambassador} The sight is dismall,\\
    And our affaires from England come too late,\\
    The eares are senselesse that should giue vs hearing,\\
    To tell him his command'ment is fulfill'd,\\
    That \rosin and \guil are dead:\\
    Where should we haue our thankes?
    
    \horspeaks Not from his mouth,\\
    Had it th'abilitie of life to thanke you:\\
    He neuer gaue command'ment for their death.\\
    But since so iumpe vpon this bloodie question,\\
    You from the Polake warres, and you from England\\
    Are heere arriued. Giue order that these bodies\\
    High on a stage be placed to the view,\\
    And let me spcake to th'yet vnknowing world,\\
    How these things came about. So shall you heare\\
    Of carnall, bloudie, and vnnaturall acts,\\
    Of accidentall iudgements, casuall slaughters\\
    Of death's put on by cunning, and forc'd cause,\\
    And in this vp shot, purposes mistooke,\\
    Falne on the Inuentors heads. All this can I\\
    Truly deliuer.
    
    \forspeaks Let vs hast to heare it,\\
    And call the Noblest to the Audience.\\
    For me, with sorrow, I embrace my Fortune,\\
    I haue some Rites of memory in this Kingdome,\\
    Which are ro claime, my vantage doth\\
    Inuite me,
    
    \horspeaks Of that I shall haue alwayes cause to speake,\\
    And from his mouth\\
    Whose voyce will draw on more:\\
    But let this same be presently perform'd,\\
    Euen whiles mens mindes are wilde,\\
    Lest more mischance\\
    On plots, and errors happen.
    
    \forspeaks Let foure Captaines\\
    Beare \ham like a Soldier to the stage,\\
    For he was likely, had he beene put on\\
    To haue prou'd most royally:\\
    And for his passage,\\
    The Souldiours Musicke, and the rites of Warre\\
    Speake lowdly for him.\\
    Take vp the body; Such a fight as this\\
    Becomes the field, but heere shewes much amis.\\
    Go, bid the Souldiers shoote.
    
    \direct{Exeunt Marching: after the which, a Peale of Ordenance are shot off.}
\end{drama*}

\end{document}